---
title: "CutTagv4R179CvsWT"
author: "Yu Kuan Pan"
date: "2026-03-19"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ROCR)
library(dplyr)
library(stringr)
library(ggplot2)
library(viridis)
library(GenomicRanges)
library(chromVAR) ## For FRiP analysis and differential analysis
library(DESeq2) ## for differential analysis section
library(ggpubr) ## for customizing figures
library(corrplot) ## for correlation plot
source("./cutTagFunction.R")
library(DT) ## for interactive tables in RMarkdown


```

if the graph below doesn't show as expect 
adjust the order here in sampleList 
@param sampleListW: List read all 3 alignment names 
        (   wild type 1% replicate 1: KWA-62277_S34_L001,
            tgf replicate 1: KWA-62278_s35_L001, 
            R179C replicate 1: KWA-62279_S36_L001)
@param histListW: name listed for gene
```{r}
sampleListW <- c("wildType_Rep1", "wildType_Rep2", "wildType_Rep3",
                 "tgf_Rep1", "tgf_Rep2", "tgf_Rep3",
                 "R179C_Rep1", "R179C_Rep2", "R179C_Rep3")
histListW <- c("wildType", "tgf", "R179C")
```

```{r}
alignResultsW <- c()
files = list.files(path = "../alignment/samU/bowtie2_summary/", pattern = "_bowtie2.txt") # adjusted based on the file destination 

files

```

### Sequencing depth 
```{r}
# Initialize an empty data frame before the loop
alignResults <- data.frame(Histone = character(),
                           Replicate = character(),
                           SequencingDepth = numeric(),
                           MappedFragNum_mm10 = numeric(),
                           AlignmentRate_mm10 = numeric(),
                           stringsAsFactors = FALSE)

# Loop through histone files
for(hist in files) {
  
  # adjusted based on the file destination
  alignRes <- read.table(paste0("../alignment/samU/bowtie2_summary/", hist), header = FALSE, fill = TRUE, stringsAsFactors = FALSE) 
  
  # Extract alignment rate and remove the '%' symbol
  alignRate <- as.numeric(substr(alignRes$V1[6], 1, nchar(as.character(alignRes$V1[6])) - 1))
  
  # Ensure the numeric conversion does not fail
  mappedFragNum <- sum(as.numeric(alignRes$V1[4]), as.numeric(alignRes$V1[5]), na.rm = TRUE)
  
  # Extract histone and replicate information from filename
  histInfo <- unlist(strsplit(hist, "_"))
  
  # Append results
  alignResults <- rbind(alignResults, data.frame(
    Histone = histInfo[1],
    Replicate = histInfo[2],
    SequencingDepth = as.numeric(as.character(alignRes$V1[1])),
    MappedFragNum_mm10 = mappedFragNum,
    AlignmentRate_mm10 = alignRate
  ))
}

alignResults %>% mutate(AlignmentRate_mm10 = paste0(AlignmentRate_mm10, "%"))

# Display the result

```

reorder the setting for visualization 
```{r}
# Define the desired order for the Histone column
desired_order <- c("wildType", "tgf", "R179C")

# Convert Histone to a factor with the specified order
alignResults$Histone <- factor(alignResults$Histone, levels = desired_order)

# Reorder the dataframe based on the factor levels
alignResults <- alignResults[order(alignResults$Histone), ]

# Display the result
print(alignResults)

```

Since we only need R179C vs WT for the poster, we can subset the data for better visualization
```{r}
alignResults <- subset(alignResults, Histone %in% c("wildType", "R179C"))
```

### Visualization for poster: Alignment Data 
```{r}
# install.packages("DT", type = "source")
library(DT)

# Create a polished table
datatable(alignResults, 
          rownames = FALSE,
          colnames = c("Condition", "Rep", "Seq Depth", "Mapped Frags", "Alignment %"),
          options = list(
              dom = 't',               # Table only (no search box)
              pageLength = 9,          # Show all rows
              columnDefs = list(list(className = 'dt-center', targets = "_all"))
          )) %>%
    # Add commas to big numbers
    formatCurrency(c('SequencingDepth', 'MappedFragNum_mm10'), currency = "", interval = 3, mark = ",", digits = 0) %>%
    # Add a heatmap effect to Alignment Rate to show quality
    formatStyle(
        'AlignmentRate_mm10',
        backgroundColor = styleInterval(c(75, 80), c('#DEEBF7', '#9ECAE1', '#4292C6')),
        fontWeight = 'bold'
)
```


##Visualizing the sequening depth and alignment results 

```{r}
fig3A = alignResults %>% ggplot(aes(x= Histone, y = SequencingDepth/1000000, fill = Histone))+geom_boxplot()+geom_jitter(aes(color= Replicate), position= position_jitter(0.15))+scale_fill_viridis(discrete = TRUE, begin = 0.1, end = 0.9, option = "magma", alpha = 0.8)+scale_color_viridis(discrete = T, begin = 0.1, end = 0.9)+theme_bw(base_size = 10)+ylab("Sequencing Depth per Million")+xlab("")+ggtitle("A. Sequencing Depth")

fig3B = alignResults %>% ggplot(aes(x= Histone, y= MappedFragNum_mm10/1000000, fill = Histone))+geom_boxplot()+geom_jitter(aes(color = Replicate), position = position_jitter(0.15)) + scale_fill_viridis(discrete = T, begin = 0.1, end = 0.9, option = "magma", alpha = 0.8) + scale_color_viridis(discrete = T, begin = 0.1, end = 0.9)+theme_bw(base_size = 10)+ylab("Mapped Fragments per Million")+xlab("")+ggtitle("B. Alignable Fragment (mm10)")

fig3C = alignResults %>% ggplot(aes(x = Histone, y = AlignmentRate_mm10, fill = Histone)) +
    geom_boxplot() +
    geom_jitter(aes(color = Replicate), position = position_jitter(0.15)) +
    scale_fill_viridis(discrete = TRUE, begin = 0.1, end = 0.9, option = "magma", alpha = 0.8) +
    scale_color_viridis(discrete = TRUE, begin = 0.1, end = 0.9) +
    theme_bw(base_size = 10) +
    ylab("% of Mapped Fragments") +
    xlab("") +
    ggtitle("C. Alignment Rate (mm10)")



ggarrange(fig3A, fig3B, fig3C, ncol = 2, nrow = 2, common.legend = T, legend = "bottom")

```

### Remove Duplicate 
```{r}
DupResults <- c()
files = list.files(path = "../alignment/removeDuplicateU/picard_summary/", pattern = "_picard.rmDup.txt")

files

```
add the duplication rate and estimated library size to the alignment summary table for better visualization 
```{r}
histList<- c("wildType", "tgf", "R179C")
# Initialize an empty data frame before the loop
DupResults <- data.frame(Histone = character(),
                         Replicate = character(),
                         #SequencingDepth = numeric(),
                         MappedFragNum_mm10 = numeric(),
                         DuplicationRateRate_mm10 = numeric(),
                         EstimatedLibrarySize= numeric(),
                         stringsAsFactors = FALSE)

# Loop through histone files
for(hist in files) {
  
  dupRes <- read.table(paste0("../alignment/removeDuplicateU/picard_summary/", hist), header = TRUE, fill = TRUE, stringsAsFactors = FALSE)
  
  histInfo<- strsplit(hist, "_")[[1]]
  DupResults<- data.frame(Histone = histInfo[1], Replicate=  histInfo[2],
                          MappedFragNum_mm10 = dupRes$READ_PAIRS_EXAMINED[1] %>% as.character %>% as.numeric, 
                          DuplicationRate_mm10 = dupRes$PERCENT_DUPLICATION[1] %>% as.character %>% as.numeric*100, 
                          EstimatedLibrarySize = dupRes$ESTIMATED_LIBRARY_SIZE[1] %>% as.character %>% as.numeric) %>% 
    mutate(UniqueFragNum = MappedFragNum_mm10 * (1-DuplicationRate_mm10/100)) %>% rbind(DupResults, .)
}

DupResults$Histone = factor(DupResults$Histone, levels = histList)
alignDupSummary = left_join(alignResults, DupResults, by = c("Histone", "Replicate", "MappedFragNum_mm10")) %>% 
  mutate(DuplicationRate_mm10 = paste0(DuplicationRate_mm10, "%"))

DupResults
DupResults<- subset(DupResults, Histone %in% c("wildType", "R179C")) # since we only need the information of R179C vs WT for the poster, we can subset the data for better visualization
alignDupSummary


# Display the result

```

```{r}
# 1. Pre-process: Handle the specific column name 'DuplicationRateRate_mm10'
# We create hidden numeric columns (Dup_Num, Align_Num) to drive the colors
plot_data <- alignDupSummary %>%
  mutate(
    # Use the exact name from your dataframe to avoid 'object not found' error
    Dup_Num = as.numeric(gsub("%", "", DuplicationRate_mm10)),
    Align_Num = as.numeric(AlignmentRate_mm10)
  )

# 2. Create the polished interactive table for the poster
datatable(plot_data, 
          rownames = FALSE,
          # Rename columns for the final display to look professional
          colnames = c("Condition", "Replicate", "Seq Depth", "Mapped Frags", 
                       "Alignment %", "Duplication %", "Est. Lib Size", "Unique Frags", 
                       "Dup_Hidden", "Align_Hidden"),
          options = list(
              dom = 't',                         # Table only (cleaner for posters)
              pageLength = 6,                    # Display all rows
              columnDefs = list(
                  list(visible = FALSE, targets = c(8, 9)), # Hide helper numeric columns
                  list(className = 'dt-center', targets = "_all")
              )
          )) %>%
    # Format integers with commas for high-quality scientific reporting
    formatCurrency(c('SequencingDepth', 'MappedFragNum_mm10', 'EstimatedLibrarySize', 'UniqueFragNum'), 
                   currency = "", interval = 3, mark = ",", digits = 0) %>%
    # Apply Blue Heatmap to Alignment Rate
    formatStyle(
        'AlignmentRate_mm10',
        'Align_Num', 
        backgroundColor = styleInterval(c(70, 75), c('#DEEBF7', '#9ECAE1', '#4292C6')),
        fontWeight = 'bold'
    ) %>%
    # Apply Green Heatmap to Duplication Rate
    formatStyle(
        'DuplicationRate_mm10',
        'Dup_Num',
        backgroundColor = styleInterval(c(50, 65), c('#F7FCF0', '#E0F3DB', '#A8DDB5'))
    )
```

### remove duplicated plot 
```{r ,fig.width = 30, fig.height = 8}
fig4A = DupResults %>% ggplot(aes(x= Histone, y= DuplicationRate_mm10, fill= Histone)) +
  geom_boxplot()+ 
  geom_jitter(aes(color = Replicate), position = position_jitter(0.15))+
  scale_fill_viridis(discrete = TRUE, begin = 0.1, end = 0.9, option = "magma", alpha = 0.8)+
  scale_color_viridis(discrete = TRUE, begin = 0.1, end = 0.9)+
  theme_bw(base_size = 14)+
  ylab("Duplication Rate (*100%)")+
  xlab("")

fig4B = DupResults %>% ggplot(aes(x= Histone, y = EstimatedLibrarySize, fill= Histone)) + 
  geom_boxplot()+
  geom_jitter(aes(color = Replicate), position = position_jitter(0.15))+
  scale_fill_viridis(discrete = TRUE, begin = 0.1, end = 0.9, option = "magma", alpha = 0.8)+
  scale_color_viridis(discrete = TRUE, begin = 0.1, end = 0.9)+
  theme_bw(base_size = 14)+
  ylab("Estimated Library Size")+
  xlab("")

fig4C = DupResults %>% ggplot(aes(x= Histone, y= UniqueFragNum, fill = Histone))+
  geom_boxplot()+
  geom_jitter(aes(color= Replicate), position = position_jitter(0.15))+
  scale_fill_viridis(discrete = TRUE, begin = 0.1, end = 0.9, option = "magma", alpha = 0.8)+
  scale_color_viridis(discrete = TRUE, begin = 0.1, end = 0.9)+
  theme_bw(base_size = 14)+
  ylab("# of Unique Fragments")+
  xlab("")

ggarrange(fig4A, fig4B, fig4C, ncol = 3, common.legend = T, legend = "bottom")
```

#### fraglen info
```{r}
FragResults <- c()
files = list.files(path = "../alignment/samU/fragmentLen/", pattern = "_fragmentLen.txt")

files

```

```{r}
fragLen <- NULL
for(hist in files){
  histInfo<- strsplit(hist, "_")[[1]]
  temp<- read.table(paste0("../alignment/samU/fragmentLen/", hist), header = F) %>% 
    mutate(fragLen = V1 %>% as.numeric, 
           fragCount = V2 %>% as.numeric, 
           Weight= as.numeric(V2)/sum(as.numeric(V2)),
           Histone = histInfo[1], 
           Replicate = histInfo[2], 
           sampleIfo = 
             paste0(histInfo[1], "_", histInfo[2])) 
    fragLen <- bind_rows(fragLen, temp)
}

fragLen<- subset(fragLen, Histone %in% c("wildType", "R179C")) # since we only need the information of R179C vs WT for the poster, we can subset the data for better visualization
```

```{r, fig.width = 12, fig.height = 4}
fragLen <- NULL

for (hist in files) {
  histInfo <- strsplit(hist, "_")[[1]]
  histoneName <- histInfo[1]  
  replicateName <- histInfo[2] 
  
  temp <- read.table(paste0("../alignment/samU/fragmentLen/", hist), header = FALSE) %>%
    mutate(
      fragLen = as.numeric(V1),
      fragCount = as.numeric(V2),
      Weight = fragCount / sum(fragCount),
      Histone = histoneName,
      Replicate = replicateName,
      sampleInfo = paste0(histInfo[1], "_", histInfo[2])
    )
  temp<- subset(temp, Histone %in% c("wildType", "R179C")) # since we only need the information of R179C vs WT for the poster, we can subset the data for better visualization
  fragLen <- bind_rows(fragLen, temp)
}

# Convert to factors
fragLen$sampleInfo <- factor(fragLen$sampleInfo, levels = sampleListW)
fragLen$Histone <- factor(fragLen$Histone, levels = c("wildType", "R179C"))  

fig5A <- fragLen %>%
  ggplot(aes(
    x = sampleInfo,
    y = fragLen,
    weight = Weight,
    fill = Histone
  )) +
  geom_violin(bw = 5) +
  scale_y_continuous(breaks = seq(0, 800, 50)) +
  scale_fill_viridis(
    discrete = TRUE,
    begin = 0.1, end = 0.9,
    option = "magma", alpha = 0.8
  ) +
  theme_bw(base_size = 12) +
  ggpubr::rotate_x_text(angle = 20) +
  ylab("Fragment Length") +
  xlab("")+ theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )


# Line plot of fragment length distributions by sample
fig5B <- fragLen %>%
  filter(!is.na(fragLen), !is.na(fragCount)) %>%
  ggplot(aes(
    x = fragLen,
    y = fragCount,
    color = Histone,
    group = sampleInfo,
    linetype = Replicate
  )) +
  geom_line(linewidth = 1) +
  scale_color_viridis(
    discrete = TRUE,
    begin = 0.1, end = 0.9,
    option = "magma"
  ) +
  theme_bw(base_size = 12) +
  xlab("Fragment Length") +
  ylab("Count") +
  coord_cartesian(xlim = c(0, 500))

# Combine the two plots side by side
ggarrange(fig5A, fig5B, nrow = 2)
```

```{r}
# 1. Rename the 'Histone' levels in your data
fragLen <- fragLen %>%
  mutate(Histone = recode(Histone, "wildType" = "WT"))

# 2. Update Figure 5A
fig5A <- fragLen %>%
  ggplot(aes(
    x = sampleInfo,
    y = fragLen,
    weight = Weight,
    fill = Histone # This will now show "WT" in the legend
  )) +
  geom_violin(bw = 5) +
  scale_y_continuous(breaks = seq(0, 800, 50)) +
  scale_fill_viridis(
    discrete = TRUE,
    begin = 0.1, end = 0.9,
    option = "magma", alpha = 0.8
  ) +
  theme_bw(base_size = 12) +
  ylab("Fragment Length") +
  xlab("") + 
  # Keeps the bottom axis clean as requested before
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.title = element_blank() # Optional: removes "Histone" title for a cleaner legend
  )

# Line plot of fragment length distributions by sample
fig5B <- fragLen %>%
  filter(!is.na(fragLen), !is.na(fragCount)) %>%
  ggplot(aes(
    x = fragLen,
    y = fragCount,
    color = Histone,
    group = sampleInfo,
    linetype = Replicate
  )) +
  geom_line(linewidth = 1) +
  scale_color_viridis(
    discrete = TRUE,
    begin = 0.1, end = 0.9,
    option = "magma"
  ) +
  theme_bw(base_size = 12) +
  xlab("Fragment Length") +
  ylab("Count") +
  coord_cartesian(xlim = c(0, 500))

# Combine the two plots side by side
ggarrange(fig5A, fig5B, nrow = 2)
```


# Update part 
the update part start from here with the new bam file and bed
## Assess Replicate reproducibility
```{r}
FragResults <- c()
files = list.files(path = "../alignment/bedWOEcoli/", pattern = "_bowtie2.fragmentsCount.bin500.bed")

files
```
```{r}
library(dplyr)
library(corrplot)

# Your file list (already loaded)
# Make sure the `files` vector is defined

# Initialize
fragCount <- NULL
 
for (file in files) {
  
  # Extract sample name (e.g., "R179C_Rep1") from the filename
  sample_name <- strsplit(file, "_KWA")[[1]][1]
  
  # Read fragment count file (3-column BED-like: chrom, bin, count)
  frag_tmp <- read.table(paste0("../alignment/bedWOEcoli/", file), header = FALSE)
  colnames(frag_tmp) <- c("chrom", "bin", sample_name)
  
  # Merge
  if (is.null(fragCount)) {
    fragCount <- frag_tmp
  } else {
    fragCount <- full_join(fragCount, frag_tmp, by = c("chrom", "bin"))
  }
}


# Define proper sample order by group
sample_order <- c(
  "R179C_Rep1", "R179C_Rep2", "R179C_Rep3",
  "wildType_Rep1", "wildType_Rep2", "wildType_Rep3"
)

# Check if all sample names were correctly extracted
if (!all(sample_order %in% colnames(fragCount))) {
  stop("Some expected sample names are missing. Please double-check filename parsing.")
}

# Select and reorder samples, log2 transform
frag_mat <- fragCount %>%
  select(all_of(sample_order)) %>%
  log2()

# Compute correlation matrix
cor_mat <- cor(frag_mat, use = "complete.obs")

```

```{r, fig.with = 10, fig.height = 12}
# Plot
corrplot(cor_mat[sample_order, sample_order],
         method = "color",
         outline = TRUE,
         addgrid.col = "darkgray",
         order = "original",   # critical to preserve your group ordering
         addrect = 3,
         rect.col = "black", rect.lwd = 3,
         cl.pos = "b", tl.col = "indianred4", tl.cex = 1,
         cl.cex = 1, addCoef.col = "black", 
         number.digits = 2, number.cex = 1,
         col = colorRampPalette(c("midnightblue", "white", "darkred"))(100))
```

```{r}
frag_mat[is.na(frag_mat)] <- 0
frag_mat_filtered <- frag_mat[apply(frag_mat, 1, function(x) var(x) > 0), ]

```

```{r}
# After computing PCA as before
pca <- prcomp(t(log2(frag_mat_filtered + 1)), scale. = TRUE)

# Assign group colors
group_colors <- rep(c("red", "green"), each = 3)

# Define sample names (same order as in frag_mat)
sample_labels <- colnames(frag_mat_filtered)

# Plot with labels
plot(pca$x[,1], pca$x[,2],
     col = group_colors,
     pch = 19, xlab = "PC1", ylab = "PC2", main = "PCA of Fragment Counts", ylim = c(-600, 2200))

# Add text labels to each point
text(pca$x[,1], pca$x[,2],
     labels = sample_labels,
     pos = 3, cex = 0.8, col = group_colors)

# Add legend
legend("topleft",
       legend = c("R179C", "wildType"),
       col = c("red", "green"),
       pch = 19)

```

```{r}
# Assuming frag_mat is already log2-transformed and ordered
frag_mat[is.na(frag_mat)] <- 0

# Filter rows with variance > 0 (as in your PCA)
frag_mat_filtered <- frag_mat[apply(frag_mat, 1, function(x) var(x) > 0), ]

# Transpose to cluster **samples**, not bins
frag_mat_t <- t(frag_mat_filtered)

# Compute distance matrix
dist_mat <- dist(frag_mat_t, method = "euclidean")

# Hierarchical clustering
hc <- hclust(dist_mat, method = "complete")  # or "average", "ward.D2", etc.

# Plot dendrogram
plot(hc,
     main = "Hierarchical Clustering of Fragment Counts",
     xlab = "",
     sub = "",
     cex = 0.9,
     hang = -1)

```

# PeakCalling 

instead of using SEACR or Scaling Factor for normalization we utilized goPeak to call peaks and then use the called peaks to do the downstream analysis.

##  GoPeak Result (slide 50 step 20)
this study apply two different argument to calculated peaks 
the first filter is slide 50, which means the window size for calculating the local background is 50bp, and the second filter is step 20, which means the step size for sliding the window is 20bp. The combination of these two parameters allows us to identify peaks with a certain level of resolution and sensitivity, while also controlling for local background noise in our ChIP-seq data.
```{r}
histL = c("R179C", "tgfB", "wildType") 
repL = paste0("Rep", 1:3)
peakSummary = data.frame()

for(hist in histL){
  # Temporary list to store GRanges for this condition
  temp_rep_list <- list()
  
  # Step 1: Read all available replicates first
  for(rep in repL){
    pattern_string <- paste0(hist, "_", rep, ".*_peaks")
    
    target_file <- list.files(
      path = "../cutTagRes/cutAndTag/goPeakRes_Step50_slide20/",
      pattern = pattern_string,
      full.names = TRUE)
    
    target_file <- target_file[!grepl("Identifier", target_file)]
    
    if(length(target_file) > 0){
      peakInfo = read.table(target_file[1], header = FALSE)
      message("Successfully read: ", basename(target_file[1]))
      
      peak_gr = GRanges(seqnames = peakInfo$V1,
                        ranges = IRanges(start = peakInfo$V2, end = peakInfo$V3))
      
      # Store the GRanges in our list named by the Replicate
      temp_rep_list[[rep]] <- peak_gr
    } else {
      warning("Could not find file for ", hist, " ", rep)
    }
  }
  
  # Step 2: Calculate the Common Peak set (Intersection of all found replicates)
  if(length(temp_rep_list) >= 2){
    # Start with the first replicate
    common_gr = temp_rep_list[[1]]
    
    # Intersect with every subsequent replicate found
    for(i in 2:length(temp_rep_list)){
      hits = findOverlaps(common_gr, temp_rep_list[[i]])
      common_gr = common_gr[unique(queryHits(hits))]
    }
    
    # This is your final consensus count (e.g., 604)
    final_consensus_count = length(common_gr)
    
    # Step 3: Create rows for the table using the final consensus count
    for(rep_name in names(temp_rep_list)){
      temp_row = data.frame(
        Histone = hist,
        Replicate = rep_name,
        peakN = length(temp_rep_list[[rep_name]]), # Total in this file (e.g., 1973)
        peakReprodNum = final_consensus_count      # Always the common count (e.g., 604)
      )
      peakSummary = rbind(peakSummary, temp_row)
    }
    
    # Optional: Save the consensus BED file for your Heatmap
    # rtracklayer::export(common_gr, paste0(hist, "_consensus_peaks.bed"))
    
  } else {
    message("Not enough replicates to calculate overlap for ", hist)
  }
}

# Step 4: Final Calculation
peakReprod = peakSummary %>% 
  mutate(peakReprodRate = (peakReprodNum / peakN) * 100)

print(peakReprod)
```

since the study only focusing on comparing the different between R179C and wildtype, we can subset the data for better visualization
```{r}
peakReprodSub<- subset(peakReprod, Histone %in% c("R179C", "wildType"))
peakReprodSub
```

```{r}
library(DT)
library(dplyr)
# 1. Pre-process: No string conversion needed since these are already integers/doubles
# We use peakReprodRate directly for the heatmap effect
plot_peak_data <- peakReprodSub 

# 2. Create the polished Peak Reproduction table
datatable(plot_peak_data, 
          rownames = FALSE,
          # Rename headers to be professional for the poster
          colnames = c("Condition", "Replicate", "Total Peaks", "Reproduced Peaks", "Reproduction %"),
          options = list(
              dom = 't',                         # Clean layout: table only
              pageLength = 6,                    # Show all rows
              columnDefs = list(
                  list(className = 'dt-center', targets = "_all")
              )
          )) %>%
    # Format peak counts with commas
    formatCurrency(c('peakN', 'peakReprodNum'), 
                   currency = "", interval = 3, mark = ",", digits = 0) %>%
    # Format the Rate to 2 decimal places
    formatRound('peakReprodRate', 2) %>%
    # Apply a Green Heatmap to Reproduction Rate (Higher = More Consistent)
    formatStyle(
        'peakReprodRate',
        backgroundColor = styleInterval(c(30, 50), c('#F7FCF0', '#E0F3DB', '#A8DDB5')),
        fontWeight = 'bold'
    )
```


## master peak (general 3 replicate sample )
the purpose is to get the master peak list for the heatmap, which is the union of all peaks across all samples. This will allow us to have a comprehensive set of peaks to analyze and visualize in the heatmap, ensuring that we capture all relevant regions of interest from our ChIP-seq data.
```{r}
mPeak = GRanges()
## overlap with bam file to get count 
for(hist in histL){
  for(rep in repL){
    pattern_string <- paste0(hist, "_", rep, ".*_peaks")
    
    target_file <- list.files(
      path = "../cutTagRes/cutAndTag/goPeakRes_Step50_slide20/",
      pattern = pattern_string,
      full.names = TRUE)
    
    target_file <- target_file[!grepl("Identifier", target_file)]
    
    if(length(target_file) > 0){
      peakInfo = read.table(target_file[1], header = FALSE)
      message("Successfully read: ", basename(target_file[1]))
      
      mPeak = GRanges(seqnames = peakInfo$V1,
                        ranges = IRanges(start = peakInfo$V2, end = peakInfo$V3), strand = "*") %>% append (mPeak, .)
    } else {
      warning("Could not find file for ", hist, " ", rep)
    }
    
    
  }
}

masterPeak = reduce(mPeak)
```

## Get the fragment counts for each peak in the master peak list 

```{r}
#library(DESeq2)

# Initialize the matrix
countMat = matrix(NA, length(masterPeak),
                  length(histL) * length(repL))
directory = "../alignment/bamWOEcoli/noEcoli/"
# Get the fragment counts for each peak in the master peak list.
i = 1
for(hist in histL){
  for(rep in repL){
    
    # 1. Create a pattern that looks for BOTH the condition and the replicate
    # Example: "R179C_Rep1" followed by anything, ending in our BAM suffix
    file_pattern <- paste0(hist, "_", rep, ".*_bowtie2\\.mapped\\.noEcoli\\.bam$")
    
    # 2. Search for the file in your directory
    target_file <- list.files(path = directory, pattern = file_pattern, full.names = TRUE)
    
    # 3. Safety check: Make sure we found exactly one file
    if(length(target_file) > 0){
      message("Successfully matched: ", basename(target_file[1]))
      
      # Use the matched filename for getCounts
      fragment_counts <- getCounts(target_file[1], masterPeak, paired = TRUE, by_rg = FALSE, format = "bam")
      countMat[, i] = counts(fragment_counts)[,1]
      
    } else {
      warning("ERROR: Could not find a file matching pattern: ", file_pattern)
    }
    
    i = i + 1
  }
}

colnames(countMat) = paste(rep(histL, each = 3), rep(repL, 3),
                           sep = "_")
```

run the code below is any of the code below run into undefine function issue

```{r}
peak_ids <- paste(seqnames(masterPeak), start(masterPeak), end(masterPeak), sep="_")
rownames(countMat) <- peak_ids
```

## Sequencing depth normalization and differential enriched peaks detection 

```{r}

selectR = which(rowSums((countMat> 0)) >= 5) # remove low count genes 
dataS = countMat[selectR,]

condition <- factor(rep(histL, each = length(repL)))

dds<- DESeqDataSetFromMatrix(countData = dataS,
                              colData = DataFrame(condition),
                              design = ~ condition)
DDS<- DESeq(dds)
normDDS<- counts(DDS, normalized=TRUE) ## normalization with respect to the sequencing depth 
colnames(normDDS) = paste0(colnames(normDDS), "_norm")
res = results(DDS, independentFiltering = FALSE,
              altHypothesis = "greaterAbs")
countMatDiff = cbind(dataS, normDDS, res)
#head(countMatDiff)
```

## compare differntial enriched peaks between different conditions 

```{r}
res_R179C<- results(DDS, contrast = c("condition", "R179C", "wildType"), independentFiltering = FALSE, altHypothesis = "greaterAbs")
```

combining raw count, normalized count and the result of differential enriched peaks detection for each comparison with R179C mutate based
```{r}
countMatDiff_R179C = cbind(dataS, normDDS, res_R179C)
```

take the significant peaks with log2FoldChange > 1 and padj < 0.05 for each comparison 
```{r}
sig_R179C <- res_R179C[which(res_R179C$padj < 0.05 & res_R179C$log2FoldChange > 1), ] # only one peaks is found even with arugement padj< 0.05  
```

```{r}
coords<- as.data.frame(masterPeak[selectR])
results_table<- cbind(coords[, c("seqnames", "start",
                                 "end")],
                      countMatDiff_R179C)
repressive_candidate<- results_table[which(results_table$padj < 0.05 & results_table$log2FoldChange > 1), ]
rownames(repressive_candidate)
```
since we only find one peak with standard filtering method 
the study will focus on the strict unique found in unique peak information 

## Match unique/strick unique peak 

## Peak Filter (R179C)
R179C unique peak match with the result table, the function will extract the peak information and the count information for the unique peak in R179C condition amount 3 replicate, the unique peak is defined as those peak that select amount one type of replicate only, like 3 replicate
```{r}
R179C_unique<- extract_target_peak(res_table = results_table, 
                             bed_path = "../cutTagRes/cutAndTag/goPeakRes_Step50_slide20/UniqueR179C.bed", 
                             master_gr = masterPeak, 
                             select_index = selectR)
```

R179 Strict Unique peak match with the result table, the function will extract the peak information and the count information for the strict unique peak in R179C condition amount 3 replicate, the strict unique peak is defined as those peak that select amount one type of replicate only, like 3 replicate and also not found in other condition 
```{r}
R179C_strict_unique<- extract_target_peak(res_table = results_table, 
                             bed_path = "../cutTagRes/cutAndTag/goPeakRes_Step50_slide20/only_R179C_strict.bed",
                             master_gr = masterPeak, 
                             select_index = selectR)
```

## inverse Match Execution 
```{r}
#BiocManager::install(c("ChIPseeker", 
#                       "TxDb.Mmusculus.UCSC.mm10.knownGene", 
#                       "org.Mm.eg.db",
#                       "clusterProfiler"))
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
```

R179C unique peak summary and annotation
```{r}
R179C_Peak_Summary<- chipSeekFunc(peakInfo = R179C_unique)
dim(R179C_Peak_Summary)
R179C_strict_Peak_Summary<- chipSeekFunc(peakInfo = R179C_strict_unique)
dim(R179C_strict_Peak_Summary)
```

## GO Enrichment Analysis

R179C unique peak GO enrichment analysis
```{r}
GOPathFunc(annotated_df = R179C_Peak_Summary, 
           pvalueCutoff = 0.05, 
           method = "BP", 
           DB = org.Mm.eg.db)
```

R179C strict unique peak GO enrichment analysis
```{r}
GOPathFunc(annotated_df = R179C_strict_Peak_Summary, 
           pvalueCutoff = 0.05, 
           method = "BP", 
           DB = org.Mm.eg.db)
```

Since we can't really see the pattern between chip seq and bulk RNA seq, we will focus on keeping promoter gene for the analysis and see how it preformed in the differential expression result in bulk RNA

R179C 
```{r}
R179C_Promoter_Summary<- R179C_Peak_Summary %>% filter(grepl("Promoter", annotation))
dim(R179C_Promoter_Summary)
head(R179C_Promoter_Summary$SYMBOL, 10)

R179C_Strict_Promoter_Summary<- R179C_strict_Peak_Summary %>% filter(grepl("Promoter", annotation))
dim(R179C_Strict_Promoter_Summary)
head(R179C_Strict_Promoter_Summary$SYMBOL, 10)
```

## bulk RNA DEG match 

#### R179C evaluation 
first we need to check if the result we found in chip seq match the differential expression results from bulk RNA seq
```{r}
# Load the bulk RNA-seq DEGs
R179CBulk <- read.csv("./sig_R179C_vs_WT.csv" , header = TRUE)
colnames(R179CBulk)[1] <- "Gene"
colnames(R179CBulk)
```

check if the promoter gene match with DEG results 
```{r}
intersect_R179C <- intersect(R179C_Promoter_Summary$SYMBOL, R179CBulk$Gene)
intersect_R179C

R179CBulk[R179CBulk$Gene %in% intersect_R179C, ]
# since R179C label 2 as promotor with arguement -5kb - 1kb write csv to export those information as well
write.csv(R179CBulk[R179CBulk$Gene %in% intersect_R179C, ], "./R179C_Bulk_Promoter_5kb_1kb.csv", row.names = FALSE)
```

```{r}
# 1. Load the library
# install.packages("DT")
library(DT)

# 2. Prepare the small comparison data
# This pulls Prune2 and Dnah8 and rounds the values for the viewer
table_dt <- R179CBulk[R179CBulk$Gene %in% c("Prune2", "Dnah8"), ]
table_dt$baseMean <- round(table_dt$baseMean, 2)
table_dt$log2FoldChange <- round(table_dt$log2FoldChange, 2)
table_dt$padj <- formatC(table_dt$padj, format = "e", digits = 2)

# 3. Show in RStudio Viewer
datatable(table_dt, 
          caption = 'Candidate Gene Comparison: R179C vs WT',
          options = list(dom = 't', ordering = FALSE), # 't' shows just the table (no search/filter)
          rownames = FALSE) %>%
  formatStyle('Gene', 
              target = 'row', 
              backgroundColor = styleEqual('Prune2', '#ffcccc')) # Highlights Prune2 row
```


check how the bulkRNA result with +- 50 kb  
```{r}
# Define the columns you want to keep
PeakListInfo <- c("seqnames", "start", "end", "width", "annotation", "distanceToTSS", "SYMBOL")

# Filter rows for the 37 genes AND filter for the columns in PeakListInfo
R179C_Filtered_Summary <- R179C_Peak_Summary[R179C_Peak_Summary$SYMBOL %in% intersect_R179C, PeakListInfo]

R179C_Filtered_Summary[abs(R179C_Filtered_Summary$distanceToTSS) < 50000, ]
R179CBulk[R179CBulk$Gene %in% R179C_Filtered_Summary[abs(R179C_Filtered_Summary$distanceToTSS) < 50000, "SYMBOL"], ]
```

## summarized necessary file

57 gene selection 
```{r}
unique(R179C_Promoter_Summary$SYMBOL)
length(unique(R179C_Promoter_Summary$SYMBOL))
write.csv(R179C_Promoter_Summary, "./R179C_Promoter_5kb_1kb_peak_Summary.csv", row.names = FALSE)
```
## Check if the study focused gene match with the peak found in R179C

```{r}
FocusGene<- c("Acta2", "Cnn1", "Tagln", "Myh11", "Tspan2", "Itga8", "Smtn",
              "Dhx32os", "Myocd", "Mrtfa", "Lmod1")
R179C_Peak_Summary[R179C_Peak_Summary$SYMBOL %in% FocusGene, ]
```

#### Gopeak result (slide 25 step 10 pval 0.05)
the whole data is under the directory goPeakRes_Step25_slide10 --pval 0.05 
```{r}
histL = c("R179C", "tgfB", "wildType") 
repL = paste0("Rep", 1:3)
peakSummary = data.frame()

for(hist in histL){
  # Temporary list to store GRanges for this condition
  temp_rep_list <- list()
  
  # Step 1: Read all available replicates first
  for(rep in repL){
    pattern_string <- paste0(hist, "_", rep, ".*_peaks")
    
    target_file <- list.files(
      path = "../cutTagRes/cutAndTag/goPeakRes_Step25_slide10/",
      pattern = pattern_string,
      full.names = TRUE)
    
    target_file <- target_file[!grepl("Identifier", target_file)]
    
    if(length(target_file) > 0){
      peakInfo = read.table(target_file[1], header = FALSE)
      message("Successfully read: ", basename(target_file[1]))
      
      peak_gr = GRanges(seqnames = peakInfo$V1,
                        ranges = IRanges(start = peakInfo$V2, end = peakInfo$V3))
      
      # Store the GRanges in our list named by the Replicate
      temp_rep_list[[rep]] <- peak_gr
    } else {
      warning("Could not find file for ", hist, " ", rep)
    }
  }
  
  # Step 2: Calculate the Common Peak set (Intersection of all found replicates)
  if(length(temp_rep_list) >= 2){
    # Start with the first replicate
    common_gr = temp_rep_list[[1]]
    
    # Intersect with every subsequent replicate found
    for(i in 2:length(temp_rep_list)){
      hits = findOverlaps(common_gr, temp_rep_list[[i]])
      common_gr = common_gr[unique(queryHits(hits))]
    }
    
    # This is your final consensus count (e.g., 604)
    final_consensus_count = length(common_gr)
    
    # Step 3: Create rows for the table using the final consensus count
    for(rep_name in names(temp_rep_list)){
      temp_row = data.frame(
        Histone = hist,
        Replicate = rep_name,
        peakN = length(temp_rep_list[[rep_name]]), # Total in this file (e.g., 1973)
        peakReprodNum = final_consensus_count      # Always the common count (e.g., 604)
      )
      peakSummary = rbind(peakSummary, temp_row)
    }
    
    # Optional: Save the consensus BED file for your Heatmap
    # rtracklayer::export(common_gr, paste0(hist, "_consensus_peaks.bed"))
    
  } else {
    message("Not enough replicates to calculate overlap for ", hist)
  }
}

# Step 4: Final Calculation
peakReprod = peakSummary %>% 
  mutate(peakReprodRate = (peakReprodNum / peakN) * 100)

print(peakReprod)
```

```{r}
library(DT)
library(dplyr)
peakReprodSub<- subset(peakReprod, Histone %in% c("R179C", "wildType"))
# 1. Pre-process: No string conversion needed since these are already integers/doubles
# We use peakReprodRate directly for the heatmap effect
plot_peak_data <- peakReprodSub 

# 2. Create the polished Peak Reproduction table
datatable(plot_peak_data, 
          rownames = FALSE,
          # Rename headers to be professional for the poster
          colnames = c("Condition", "Replicate", "Total Peaks", "Reproduced Peaks", "Reproduction %"),
          options = list(
              dom = 't',                         # Clean layout: table only
              pageLength = 6,                    # Show all rows
              columnDefs = list(
                  list(className = 'dt-center', targets = "_all")
              )
          )) %>%
    # Format peak counts with commas
    formatCurrency(c('peakN', 'peakReprodNum'), 
                   currency = "", interval = 3, mark = ",", digits = 0) %>%
    # Format the Rate to 2 decimal places
    formatRound('peakReprodRate', 2) %>%
    # Apply a Green Heatmap to Reproduction Rate (Higher = More Consistent)
    formatStyle(
        'peakReprodRate',
        backgroundColor = styleInterval(c(30, 50), c('#F7FCF0', '#E0F3DB', '#A8DDB5')),
        fontWeight = 'bold'
    )
```


## master peak (general 3 replicate sample )
the purpose is to get the master peak list for the heatmap, which is the union of all peaks across all samples. This will allow us to have a comprehensive set of peaks to analyze and visualize in the heatmap, ensuring that we capture all relevant regions of interest from our ChIP-seq data.
```{r}
mPeak = GRanges()
## overlap with bam file to get count 
for(hist in histL){
  for(rep in repL){
    pattern_string <- paste0(hist, "_", rep, ".*_peaks")
    
    target_file <- list.files(
      path = "../cutTagRes/cutAndTag/goPeakRes_Step25_slide10/",
      pattern = pattern_string,
      full.names = TRUE)
    
    target_file <- target_file[!grepl("Identifier", target_file)]
    
    if(length(target_file) > 0){
      peakInfo = read.table(target_file[1], header = FALSE)
      message("Successfully read: ", basename(target_file[1]))
      
      mPeak = GRanges(seqnames = peakInfo$V1,
                        ranges = IRanges(start = peakInfo$V2, end = peakInfo$V3), strand = "*") %>% append (mPeak, .)
    } else {
      warning("Could not find file for ", hist, " ", rep)
    }
    
    
  }
}

masterPeak = reduce(mPeak)
```

## Get the fragment counts for each peak in the master peak list 

```{r}
#library(DESeq2)

# Initialize the matrix
countMat = matrix(NA, length(masterPeak),
                  length(histL) * length(repL))

# Get the fragment counts for each peak in the master peak list.
i = 1
for(hist in histL){
  for(rep in repL){
    
    # 1. Create a pattern that looks for BOTH the condition and the replicate
    # Example: "R179C_Rep1" followed by anything, ending in our BAM suffix
    file_pattern <- paste0(hist, "_", rep, ".*_bowtie2\\.mapped\\.noEcoli\\.bam$")
    
    # 2. Search for the file in your directory
    target_file <- list.files(path = "../alignment/bamWOEcoli/noEcoli/", pattern = file_pattern, full.names = TRUE)
    
    # 3. Safety check: Make sure we found exactly one file
    if(length(target_file) > 0){
      message("Successfully matched: ", basename(target_file[1]))
      
      # Use the matched filename for getCounts
      fragment_counts <- getCounts(target_file[1], masterPeak, paired = TRUE, by_rg = FALSE, format = "bam")
      countMat[, i] = counts(fragment_counts)[,1]
      
    } else {
      warning("ERROR: Could not find a file matching pattern: ", file_pattern)
    }
    
    i = i + 1
  }
}

colnames(countMat) = paste(rep(histL, each = 3), rep(repL, 3),
                           sep = "_")
```

run the code below is any of the code below run into undefine function issue

```{r}
peak_ids <- paste(seqnames(masterPeak), start(masterPeak), end(masterPeak), sep="_")
rownames(countMat) <- peak_ids
```

## Sequencing depth normalization and differential enriched peaks detection 

```{r}

selectR = which(rowSums((countMat> 0)) >= 5) # remove low count genes 
dataS = countMat[selectR,]

condition <- factor(rep(histL, each = length(repL)))

dds<- DESeqDataSetFromMatrix(countData = dataS,
                              colData = DataFrame(condition),
                              design = ~ condition)
DDS<- DESeq(dds)
normDDS<- counts(DDS, normalized=TRUE) ## normalization with respect to the sequencing depth 
colnames(normDDS) = paste0(colnames(normDDS), "_norm")
res = results(DDS, independentFiltering = FALSE,
              altHypothesis = "greaterAbs")
countMatDiff = cbind(dataS, normDDS, res)
#head(countMatDiff)
```

## compare differntial enriched peaks between different conditions 

```{r}
res_R179C<- results(DDS, contrast = c("condition", "R179C", "wildType"), independentFiltering = FALSE, altHypothesis = "greaterAbs")
```

take the significant peaks with log2FoldChange > 1 and padj < 0.05 for each comparison 
```{r}
sig_R179C <- res_R179C[which(res_R179C$padj < 0.05 & res_R179C$log2FoldChange > 1), ]
```

combining raw count, normalized count and the result of differential enriched peaks detection for each comparison with R179C mutate based
```{r}
countMatDiff_R179C = cbind(dataS, normDDS, res_R179C)

coords<- as.data.frame(masterPeak[selectR])
results_table_R179C<- cbind(coords[, c("seqnames", "start",
                                 "end")],
                      countMatDiff_R179C)
repressive_candidate_R179C<- results_table_R179C[which(results_table_R179C$padj < 0.05 & results_table_R179C$log2FoldChange > 1), ]
rownames(repressive_candidate_R179C)
```

from the result we got from repressive_candidate_R179C the only peak shown as significant is the peak located at: 
chr17: 11327880: 11329215
## Match unique/2 out of 3 unique peak 

## Peak Filter (R179C)
R179C unique peak match with the result table, the function will extract the peak information and the count information for the unique peak in R179C condition amount 3 replicate, the unique peak is defined as those peak that select amount one type of replicate only, like 3 replicate
```{r}
R179C_unique<- extract_target_peak(res_table = results_table_R179C, 
                             bed_path = "../cutTagRes/cutAndTag/goPeakRes_Step25_slide10/unique_R179C.bed", 
                             master_gr = masterPeak, 
                             select_index = selectR)
```

R179 Strict Unique peak match with the result table, the function will extract the peak information and the count information for the strict unique peak in R179C condition amount 3 replicate, the strict unique peak is defined as those peak that select amount one type of replicate only, like 3 replicate and also not found in other condition 
```{r}
R179C_2out3<- extract_target_peak(res_table = results_table_R179C, 
                             bed_path = "../cutTagRes/cutAndTag/goPeakRes_Step25_slide10/R179C_2out3.bed",
                             master_gr = masterPeak, 
                             select_index = selectR)
```

## inverse Match Execution 
```{r}
#BiocManager::install(c("ChIPseeker", 
#                       "TxDb.Mmusculus.UCSC.mm10.knownGene", 
#                       "org.Mm.eg.db",
#                       "clusterProfiler"))
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
```

R179C unique peak summary and annotation
```{r}
R179C_Peak_Summary<- chipSeekFunc(peakInfo = R179C_unique)
dim(R179C_Peak_Summary)
R179C_2out3_Peak_Summary<- chipSeekFunc(peakInfo = R179C_2out3)
dim(R179C_2out3_Peak_Summary )
```

## promotor check -5kb to +1kb

#### R179C unique peak/2out3 promotor check 
```{r}
R179C_Promoter_Summary<- R179C_Peak_Summary %>% filter(grepl("Promoter", annotation))
dim(R179C_Promoter_Summary)
head(R179C_Promoter_Summary$SYMBOL, 10)

R179C_2out3_Promoter_Summary<- R179C_2out3_Peak_Summary %>% filter(grepl("Promoter", annotation))
dim(R179C_2out3_Promoter_Summary)
head(R179C_2out3_Promoter_Summary$SYMBOL, 10)
```

## bulk RNA DEG match 

#### R179C evaluation (-5kb to +1kb)
first we need to check if the result we found in chip seq match the differential expression results from bulk RNA seq
```{r}
# Load the bulk RNA-seq DEGs
R179CBulk <- read.csv("./sig_R179C_vs_WT.csv" , header = TRUE)
colnames(R179CBulk)[1] <- "Gene"
colnames(R179CBulk)
```
check if the promoter gene match with DEG results (R179C unique)
```{r}
intersect_R179C <- intersect(R179C_Promoter_Summary$SYMBOL, R179CBulk$Gene)
intersect_R179C

R179CBulk[R179CBulk$Gene %in% intersect_R179C, ]
```

```{r}
library(DT)

# Create the first table
prune2_tab <- R179CBulk[R179CBulk$Gene %in% intersect_R179C, ]

datatable(prune2_tab, 
          caption = 'Table 1: Target Gene Validation (Prune2)',
          options = list(dom = 't'), 
          rownames = FALSE) %>%
  formatRound(columns=c('baseMean', 'log2FoldChange'), digits=2) %>%
  formatSignif(columns=c('pvalue', 'padj'), digits=3) %>%
  # Highlight Prune2 row
  formatStyle('Gene', target = 'row', 
              backgroundColor = styleEqual('Prune2', '#ffcccc'))
```


check if the promoter gene match with DEG results (R179C 2out3)
```{r}
intersect_R179C_2out3 <- intersect(R179C_2out3_Promoter_Summary$SYMBOL, R179CBulk$Gene)
intersect_R179C_2out3

R179CBulk[R179CBulk$Gene %in% intersect_R179C_2out3, ]
```

```{r}
# Create the intersection table
intersect_tab <- R179CBulk[R179CBulk$Gene %in% intersect_R179C_2out3, ]

datatable(intersect_tab, 
          caption = 'Table 2: Genes Matching Promoter Results (2/3 Replicates)',
          options = list(pageLength = 10, dom = 'tip'), 
          rownames = FALSE) %>%
  formatRound(columns=c('baseMean', 'log2FoldChange'), digits=2) %>%
  formatSignif(columns=c('pvalue', 'padj'), digits=3) %>%
  # Highlight Prune2 row within the larger list
  formatStyle('Gene', target = 'row', 
              backgroundColor = styleEqual('Prune2', '#ffcccc'))
```

