---
title: "BulkRNAv3"
author: "Yu Kuan Pan"
date: "2026-03-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SummarizedExperiment)
library(tximport)
library(readr)
library(GenomicFeatures) # for creating TxDb object
library(txdbmaker) 
library(tidyverse) # for data manipulation and visualization
library(DESeq2) # for differential expression analysis
library(edgeR) # for differential expression analysis
library(pheatmap) # for heatmap
library(readxl) # for reading excel file
library(clusterProfiler) # pathway analysis package 
library(org.Mm.eg.db) # mouse annotation package
#remotes::install_github("YuLab-SMU/ReactomePA", ref = "devel")
library(ReactomePA)
```

# install library if not already install 
un-comment this part if the packages is not installed yet 
```{r}
#if(!requireNamespace("BiocManager", quietly = T)){
#  install.packages("BiocManager")
#  BiocManager::install(version = "3.20")
#}

# Use BiocManager to install tximport 
#BiocManager::install("tximport")
#BiocManager::install("edgeR")

#BiocManager::install("GenomicFeatures")
#BiocManager::install("txdbmaker")

#if(!requireNamespace("biomaRt", quietly = T)){
#  install.packages("biomaRt")
#}

#BiocManager::install("rtracklayer")
```

## load in necessary Cell Count information  
```{r}
library(data.table)

counts_dir<- "../../bulkRNAData/"
files <- list.files(
  counts_dir,
  pattern = "\\.counts(\\.txt)?$",
  full.names = TRUE,
  ignore.case = TRUE
)
files <- files[!grepl("Zone\\.Identifier", files, ignore.case = TRUE)]


# filename -> clean sample name
pretty_name <- function(f) {
  nm <- sub("\\.counts(\\.txt)?$", "", basename(f))  # drop extension
  nm <- sub("(_S\\d+.*)$", "", nm)                   # drop _Sxx... suffix
  nm <- sub("(\\d)percent-(\\d)$", "\\1percent\\2", nm)  # 1percent-3 -> 1percent3
  nm
}

# read one file (first two cols only), name second col as sample
read_one <- function(f) {
  dt <- fread(
    f, header = FALSE, data.table = TRUE,
    select = 1:2,                      # only first 2 columns
    col.names = c("gene", "count")     # set column names directly
  )
  dt[, gene := as.character(gene)]
  dt[, count := as.integer(count)]
  # collapse duplicates if any
  dt <- dt[, .(count = sum(count, na.rm = TRUE)), by = gene]
  setnames(dt, "count", pretty_name(f))  # rename 2nd col to sample name
  dt
}

lst <- lapply(files, read_one)

# merge by gene (union) and fill missing with 0
merged <- Reduce(function(x, y) merge(x, y, by = "gene", all = TRUE), lst)
for (j in 2:ncol(merged)) {
  set(merged, which(is.na(merged[[j]])), j, 0L)
}

count_matrix <- as.matrix(merged[, -1])
rownames(count_matrix) <- merged$gene
storage.mode(count_matrix) <- "integer"

dim(count_matrix)
colnames(count_matrix)
```

```{r}
summary(count_matrix)
```

```{r}
dim(count_matrix)
```

```{r}
hist(count_matrix, main= "Gene Expression", xlab = "Counts")
```

## Identify Marker gene with Differential Expression 
```{r}
sample_info<- data.frame(
  sampleName = c("R179C-1percent-1_S21",
                 "R179C-1percent-2_S24",
                 "R179C-1percent-3_S27",
                 "WT-1percent-1_S19",
                 "WT-1percent-2_S22",
                 "WT-1percent-3_S25",
                 "WT-TGFb-1percent-1_S20",
                 "WT-TGFb-1percent-2_S23",
                 "WT-TGFb-1percent-3_S26"),
  condition = c("R179C", "R179C", "R179C",
                "WT", "WT", "WT",
                "TGFb", "TGFb", "TGFb"),
  replicate = c("1", "2", "3",
                "1", "2", "3",
                "1", "2", "3")
  
)
rownames(sample_info)<- sample_info$sampleName
```

## Data Cleaning and Preprocessing
the following chuck is designed to filter out those 
```{r}
dge_all<- DGEList(counts = count_matrix)
design_filter <- model.matrix(~condition, data= sample_info)
keep<- filterByExpr(dge_all, design = design_filter, min.cont = 10)
filtered_count_matrix<- count_matrix[keep, ]
```

```{r}
summary(filtered_count_matrix)
```

```{r}
dim(filtered_count_matrix)
```

```{r}
hist(filtered_count_matrix, main= "Gene Expression", xlab = "Counts")
```

```{r}
count_matrix<- filtered_count_matrix
```

## Exchange gene id with actual gene name 
this chunk will be checked later since some of the gene aren't able to matched with the ID use control shift c to comment all chunk
```{r}

stopifnot(all(count_matrix == round(count_matrix)))
counts<- as.matrix(count_matrix)

# strip version suffix from Ensembl IDs
ens <- sub("\\.\\d+$", "", rownames(counts))

mapfile<- "../../mm10_ens2symbol.tsv"

map<- read.delim(mapfile, header = F, sep = "\t",
                 col.names =
                   c("ensembl_gene_id", "gene_name"),
                 stringsAsFactors = F)

# check for both gene ID match ens
matched_ids <- intersect(ens, map$ensembl_gene_id)
length(matched_ids)
head(matched_ids, 10)

# Remove ".5" or ".16" etc.
ens_ids <- sub("\\.\\d+$", "", rownames(count_matrix))
rownames(count_matrix) <- ens_ids

# make sure column names in map are: ensembl_gene_id, gene_name
colnames(map) <- c("ensembl_gene_id", "gene_name")

# remove duplicated Ensembl IDs in the map
map <- map[!duplicated(map$ensembl_gene_id), ]

# create a vector for quick lookup
gene_names <- setNames(map$gene_name, map$ensembl_gene_id)

# replace Ensembl IDs by matched gene name
matched_genes <- gene_names[rownames(count_matrix)]

# any NA means not found → keep Ensembl ID
matched_genes[is.na(matched_genes)] <- rownames(count_matrix)[is.na(matched_genes)]

# assign new names
rownames(count_matrix) <- matched_genes
colnames(count_matrix)<- rownames(sample_info)
```

## creating DESeq object for Differential Expression
```{r}
colnames(count_matrix)<- rownames(sample_info)
dds<- DESeqDataSetFromMatrix(countData = count_matrix,
                             colData = sample_info,
                             design = ~ condition)

dds$condition <- relevel(dds$condition, ref = "WT")

dds<- DESeq(dds)

```

```{r}
# pairwise
res_R179C_vs_WT   <- results(dds, contrast = c("condition","R179C","WT"))
res_TGFb_vs_WT    <- results(dds, contrast = c("condition","TGFb","WT"))
```

```{r}
# global “compare all together” (ANOVA-like)
dds_lrt <- DESeq(dds, test="LRT", reduced=~1)
res_LRT <- results(dds_lrt)   # genes that vary across WT/R179C/TGFb
```

## PCA and sample distance heatmap
```{r}
set.seed(123)
# subset to only R179C and WT samples for clearer visualization
samples_to_keep<- sample_info$condition %in% c("R179C", "WT")
dds_subset <- dds[, samples_to_keep]
dds_subset$condition <- factor(dds_subset$condition, levels = c("WT", "R179C"))  # drop unused levels
vsd_subset<- vst(dds_subset, blind = FALSE)
dds<- dds_subset
vsd<- vsd_subset
mat<- assay(vsd_subset)
#vsd <- vst(dds, blind = FALSE)
#mat <- assay(vsd)

# PCA with sample labels
library(ggplot2)
library(ggrepel) # Highly recommended for PCA labels

p <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(p, "percentVar"))

ggplot(p, aes(PC1, PC2, color = condition, label = name)) +
  geom_point(size = 3) +
  # geom_text_repel handles the "off-screen" and overlapping labels automatically
  geom_text_repel(size = 3, box.padding = 0.5, max.overlaps = Inf) +
  # Expand the X-axis limits manually to ensure the long text strings fit
  expand_limits(x = c(min(p$PC1) - 5, max(p$PC1) + 10)) + 
  xlab(paste0("PC1: ", percentVar[1], "%")) +
  ylab(paste0("PC2: ", percentVar[2], "%")) +
  ggtitle("PCA (vst) — all samples") +
  theme_bw() # Professional look for posters


# Sample distance heatmap
d <- dist(t(assay(vsd)))
pheatmap(as.matrix(d), annotation_col=as.data.frame(colData(dds)[,"condition",drop=FALSE]))
```

```{r}
library(ggrepel)
# 1. Create the clean labels (WT-1, WT-2, WT-3, R179C-1...)
# This creates a sequence (1, 2, 3) for each 'condition' group
replicate_numbers <- ave(seq_along(dds_subset$condition), 
                         dds_subset$condition, 
                         FUN = seq_along)

# 2. Combine the Condition with the Number
# Result: "WT-1", "WT-2", "WT-3", "R179C-1", "R179C-2", "R179C-3"
new_names <- paste0(dds_subset$condition, "-", replicate_numbers)

# 3. Apply to the dds object
colnames(dds_subset) <- new_names

# 4. Proceed to VST and PCA
vsd_subset <- vst(dds_subset, blind = FALSE)
p <- plotPCA(vsd_subset, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(p, "percentVar"))

# 5. PCA Plot with Clean labels
ggplot(p, aes(PC1, PC2, color = condition, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(size = 4, box.padding = 0.5, fontface = "bold") +
  xlab(paste0("PC1: ", percentVar[1], "%")) +
  ylab(paste0("PC2: ", percentVar[2], "%")) +
  theme_bw(base_size = 14) +
  ggtitle("PCA: R179C vs WT ")

# 6. Sample Distance Heatmap
d <- dist(t(assay(vsd_subset)))
pheatmap(as.matrix(d), 
         clustering_distance_rows = d,
         clustering_distance_cols = d,
         annotation_col = as.data.frame(colData(dds_subset)[,"condition",drop=FALSE]),
         main = "Sample Distance Heatmap")
```


### R179C vs wildType 
```{r}
sig_R179C_vs_WT<- subset(res_R179C_vs_WT, padj < 0.05
                         & abs(log2FoldChange) > 1)
sig_R179C_vs_WT<- as.data.frame(sig_R179C_vs_WT[order(sig_R179C_vs_WT$log2FoldChange, decreasing = T),])
nrow(sig_R179C_vs_WT)
head(sig_R179C_vs_WT, 10)
```

remove the argument for log2FC > 1
```{r}
sig_pval_R179C_vs_WT<- subset(res_R179C_vs_WT, padj < 0.05)
sig_pval_R179C_vs_WT<- as.data.frame(sig_pval_R179C_vs_WT[order(sig_pval_R179C_vs_WT$log2FoldChange, decreasing = T),])
nrow(sig_pval_R179C_vs_WT)
#write.csv(sig_pval_R179C_vs_WT, "./bulkRNA3repRes/sig_pval_R179C_vs_WT.csv")
```
### validate gene list result with previous study WT vs R179C
library tidyverse include the read_excel from readxl library 
the chunk below is to validate the DEG result that we receive from WT vs R179C, in the previous study we found 
```{r}
sheets<- excel_sheets("../../bulkRNARefence1Replicate/DEGs.xlsx")
print(sheets)
R179C_WT_ref<- read_excel(
  "../../bulkRNARefence1Replicate/DEGs.xlsx",
           sheet = "1vs2") %>% 
  select(geneName, log2FC, Probability)
nrow(R179C_WT_ref)
head(R179C_WT_ref, 10)
```

#### check common gene 
```{r}
InterGene<- intersect(rownames(sig_R179C_vs_WT), R179C_WT_ref$geneName )
nrow(sig_R179C_vs_WT[InterGene,])
head(sig_R179C_vs_WT[InterGene,],10)
```

```{r}
write.csv(as.data.frame(sig_R179C_vs_WT), "./sig_R179C_vs_WT.csv")
```


```{r}
save(sig_R179C_vs_WT, file = "./sig_R179C_vs_WT.RData")
```

## Create MA & volcano plot 

#### R179C  vs WT
```{r}
DESeq2::plotMA(res_R179C_vs_WT, main = "R179C vs WT", ylim = c(-5,5),
       alpa=0.05)
```

### Volcano plot R179C vs WT
```{r}
#devtools::install_github('kevinblighe/EnhancedVolcano')
library(EnhancedVolcano)

EnhancedVolcano(res_R179C_vs_WT,
                lab = rownames(res_R179C_vs_WT),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'R179C vs WT',
                pCutoff = 0.05,
                FCcutoff = 1.0,
                pointSize = 2.0,
                labSize = 3.0)
```

```{r}
# 1. set res_R179C vs WT as data frame for easier manipulation
res_df <- as.data.frame(res_R179C_vs_WT)
res_robust <- res_df[which(res_df$baseMean > 600 & !is.na(res_df$padj)), ]

# 2. Identify Top 40 for each condition (R179C up and WT up/R179C down)
# Sorting by Log2FC for magnitude
top10_R179C <- head(rownames(res_robust[order(res_robust$log2FoldChange, decreasing = TRUE), ]), 10)
top10_WT    <- head(rownames(res_robust[order(res_robust$log2FoldChange, decreasing = FALSE), ]), 10)

# 3. Combine them and explicitly add Prune2
# unique() prevents errors if Prune2 is already in the top 40
genes_to_label <- unique(c("Prune2",top10_R179C, top10_WT))

# 4. Generate the Volcano Plot
library(EnhancedVolcano)
EnhancedVolcano(res_R179C_vs_WT,
                lab = rownames(res_R179C_vs_WT),
                selectLab = genes_to_label, # This is the key line
                x = 'log2FoldChange',
                y = 'padj',
                title = 'R179C vs WT baseMean > 600',
                pCutoff = 0.05,
                FCcutoff = 1.0,
                pointSize = 2.0,
                labSize = 7.0,
                drawConnectors = TRUE,   # Adds lines from label to point
                widthConnectors = 0.5,
                colConnectors = 'black')
```

```{r}
# 1. Identify the Top 40 robust genes per side
res_df <- as.data.frame(res_R179C_vs_WT)
res_robust <- res_df[which(res_df$baseMean > 600 & !is.na(res_df$padj)), ]

top40_up <- head(rownames(res_robust[order(res_robust$log2FoldChange, decreasing = TRUE), ]), 40)
top40_down <- head(rownames(res_robust[order(res_robust$log2FoldChange, decreasing = FALSE), ]), 40)

# 2. label list 
my_labels <- unique(c(top40_up, top40_down))

# 3. Generate the Volcano Plot (Standard Colors)
library(EnhancedVolcano)
EnhancedVolcano(res_df,
                lab = rownames(res_df),
                x = 'log2FoldChange',
                y = 'padj',
                selectLab = my_labels,
                title = 'R179C vs WT Transcriptional Profile',
                subtitle = 'Top 40 Robust DEGs (baseMean > 600)',
                pCutoff = 0.05,
                FCcutoff = 1.0,
                # Standard Colors: NS=grey, Log2FC=green, p-value=blue, both=red
                col = c('grey', 'forestgreen', 'royalblue', 'red2'),
                pointSize = 2.0,
                labSize = 3.5,
                # Ensure all 81 names are visible
                max.overlaps = Inf, 
                drawConnectors = TRUE,
                widthConnectors = 0.5,
                colConnectors = 'black',
                legendPosition = 'right',
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
```
#### Volcano plot (baseMean > 600)
```{r}
# 1. Setup - We only want ONE label
res_df <- as.data.frame(res_R179C_vs_WT)
my_label <- "Prune2"

# 2. Plotting (Minimalist for the Poster)
library(EnhancedVolcano)
EnhancedVolcano(res_df,
                lab = rownames(res_df),
                x = 'log2FoldChange',
                y = 'padj',
                selectLab = my_label,        # ONLY Prune2 will be labeled
                title = 'R179C vs WT: Transcriptional Validation',
                subtitle = 'Validated Target: Prune2 ',
                pCutoff = 0.05,
                FCcutoff = 1.0,
                # Using your standard colors
                col = c('grey', 'forestgreen', 'royalblue', 'red2'),
                pointSize = 2.0,
                labSize = 6.0,               # Made the label larger since it is the only one
                # --- CLEAN CONNECTOR ---
                drawConnectors = TRUE,
                widthConnectors = 0.6,
                colConnectors = 'black',
                arrowheads = FALSE,          # Just the straight line you wanted
                lengthConnectors = unit(0.1, "npc"), 
                # -----------------------
                max.overlaps = Inf,
                gridlines.major = FALSE,
                gridlines.minor = FALSE,
                legendPosition = 'right')    
```

## plot count example 
use Prune2 gene as example
```{r}
plotCounts(dds, gene="Prune2", intgroup="condition")
```

## Heatmap of top 20 DEGs

#### R179C vs WT
```{r}
top20_R179C_vs_WT <- head(rownames(sig_R179C_vs_WT), 20)
mat_R179C_vs_WT <- assay(vst(dds))[top20_R179C_vs_WT, ]
mat_R179C_vs_WT <- mat_R179C_vs_WT - rowMeans(mat_R179C_vs_WT)
pheatmap(mat_R179C_vs_WT, annotation_col = as.data.frame(colData(dds)[, "condition", drop=F]))
```

```{r}
# R179C upregulated genes
#r179c_up <- subset(res_R179C_vs_WT, log2FoldChange > 1 & padj < 0.05)
#top20_R179C <- head(rownames(r179c_up[order(r179c_up$padj), ]), 20)
top20_R179C <- head(rownames(sig_R179C_vs_WT), 20)

# WT upregulated genes
#wt_up <- subset(res_R179C_vs_WT, log2FoldChange < -1  & padj < 0.05)
#top20_WT <- head(rownames(wt_up[order(wt_up$padj), ]), 20)

# Already filtered for padj < 0.05 and |log2FC| > 1
sig_R179C_vs_WT_WT <- sig_R179C_vs_WT[order(sig_R179C_vs_WT$log2FoldChange), ]
top20_WT <- head(rownames(sig_R179C_vs_WT_WT), 20)

# Combine gene set
combined_genes <- c(top20_R179C, top20_WT)

# Subset samples to R179C and WT only 
selected_samples <- colData(dds)$condition %in% c("R179C", "WT")

mat_combined <- assay(vst(dds))[combined_genes, selected_samples]
mat_combined <- mat_combined - rowMeans(mat_combined)

annotation <- as.data.frame(colData(dds)[selected_samples, "condition", drop=FALSE])

pheatmap(mat_combined,
         annotation_col = annotation,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE, 
         main = "Heatmap of Top 20 DE Genes: R179C vs WT")

```

```{r}
# 1. Filter for Robustness (baseMean > 600)
# Assuming sig_R179C_vs_WT is your dataframe of significant results
sig_robust <- sig_R179C_vs_WT[sig_R179C_vs_WT$baseMean > 600, ]

# 2. Extract Top 40 R179C Enriched (Highest Log2FC)
# We sort by log2FoldChange decreasing to get the most upregulated
top40_R179C <- head(rownames(sig_robust[order(sig_robust$log2FoldChange, decreasing = TRUE), ]), 40)

# 3. Extract Top 40 WT Enriched (Lowest Log2FC / Most Repressed)
# We sort by log2FoldChange increasing to get the most downregulated
top40_WT <- head(rownames(sig_robust[order(sig_robust$log2FoldChange, decreasing = FALSE), ]), 40)

# 4. Combine and Ensure Prune2 is included
# Since Prune2 is around rank 40, we add it explicitly to be safe
combined_genes <- unique(c(top40_R179C, top40_WT))

# 5. Subset samples correctly (Ensure 3 replicates only)
# We filter for condition AND ensure we only take replicates 1, 2, and 3
selected_samples <- colData(dds)$condition %in% c("R179C", "WT") & 
                    colData(dds)$replicate %in% c("1", "2", "3")

# 6. Prepare Matrix (Using VST values)
# We use the subset of samples we just defined
vst_mat <- assay(vst(dds))
mat_combined <- vst_mat[combined_genes, selected_samples]

# Row-center the data (Z-score style) to highlight differences
mat_combined <- mat_combined - rowMeans(mat_combined)

# 7. Create Heatmap
annotation <- as.data.frame(colData(dds)[selected_samples, "condition", drop=FALSE])

pheatmap(mat_combined,
         annotation_col = annotation,
         cluster_rows = FALSE,
         cluster_cols = TRUE,
         show_rownames = TRUE, 
         fontsize_row = 6, # Smaller font since we have 80+ genes
         main = "Heatmap: Top 40 Robust DE Genes (baseMean > 600)")


```

```{r}
# Load required libraries
library(pheatmap)
library(grid)

# 1. Filter for Robustness (baseMean > 600)
sig_robust <- sig_R179C_vs_WT[sig_R179C_vs_WT$baseMean > 600, ]

# 2. Extract Top 40 R179C Enriched (Highest Log2FC)
top40_R179C <- head(rownames(sig_robust[order(sig_robust$log2FoldChange, decreasing = TRUE), ]), 40)

# 3. Extract Top 40 WT Enriched (Lowest Log2FC)
top40_WT <- head(rownames(sig_robust[order(sig_robust$log2FoldChange, decreasing = FALSE), ]), 40)

# 4. Combine and Ensure Prune2 is included
# Using unique() prevents duplicates if Prune2 is already in the top 40
combined_genes <- unique(c(top40_R179C, top40_WT))

# 5. Subset samples (R179C and WT, Replicates 1-3)
selected_samples <- colData(dds)$condition %in% c("R179C", "WT") & 
                    colData(dds)$replicate %in% c("1", "2", "3")

# 6. Prepare Matrix (Using VST values)
vst_mat <- assay(vst(dds))
mat_combined <- vst_mat[combined_genes, selected_samples]

# --- IMPROVED SCALING ---
# Instead of just centering, we use Z-scores so all genes have equal visual weight
mat_scaled <- t(scale(t(mat_combined)))

# 7. Create Annotation Dataframe
annotation <- as.data.frame(colData(dds)[selected_samples, "condition", drop=FALSE])

# 8. Generate the Heatmap Object (Store it in 'p' instead of drawing immediately)
p <- pheatmap(mat_scaled,
              annotation_col = annotation,
              cluster_rows = FALSE,   # Keep your Top 40 / Bottom 40 order
              cluster_cols = TRUE,    # Cluster samples to see groups
              show_rownames = TRUE, 
              fontsize_row = 5,       # Slightly larger for readability
              main = "Heatmap: Top 40 Robust DE Genes (BaseMean > 600)")

# 9. INJECT CUSTOM COLORS FOR PRUNE2
# Find which grob (graphical object) contains the row names
cols <- rep("black", nrow(mat_scaled))
# Find the exact position of Prune2 in the matrix rows
prune2_idx <- which(rownames(mat_scaled) == "Prune2")

# If Prune2 exists in your list, color it red and make it bold
if(length(prune2_idx) > 0){
  cols[prune2_idx] <- "red"
  
  # Get the row names grob
  row_names_grob <- p$gtable$grobs[[which(p$gtable$layout$name == "row_names")]]
  
  # Apply the color vector and bold face
  row_names_grob$gp$col <- cols
  row_names_grob$gp$fontface <- ifelse(cols == "red", "bold", "plain")
  
  # Put the modified grob back into the plot object
  p$gtable$grobs[[which(p$gtable$layout$name == "row_names")]] <- row_names_grob
}

# 10. Draw the final Result
grid.newpage()
grid.draw(p$gtable)
```


## pathway analysis function
```{r}
suppressMessages({
  library(clusterProfiler)
  library(ReactomePA)
  library(org.Mm.eg.db)
  library(ggplot2)
  library(openxlsx)
})

Pathway_Enrich <- function(sig_gene_list,
                           condition = "Sample",
                           method = c("KEGG", "Reactome"),
                           p_cut = 0.05,
                           showCategory = 10,
                           output_prefix = "Pathway") {
  
  method <- match.arg(method)  # restrict to KEGG or Reactome
  
  message(paste0("Running ", method, " enrichment for ", condition, " ..."))
  
  # ---- Convert gene symbols to Entrez IDs
  gene_symbols <- rownames(sig_gene_list)
  entrez_ids <- suppressMessages(
    bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
  )
  
  if (is.null(entrez_ids) || nrow(entrez_ids) == 0) {
    warning("No valid Entrez IDs found for enrichment.")
    return(NULL)
  }
  
  # ---- Run enrichment analysis
  enrich_res <- switch(
    method,
    KEGG = suppressMessages(
      enrichKEGG(gene = entrez_ids$ENTREZID,
                 organism = "mmu",
                 pvalueCutoff = p_cut)
    ),
    Reactome = suppressMessages(
      enrichPathway(gene = entrez_ids$ENTREZID,
                    organism = "mouse",
                    pvalueCutoff = p_cut)
    )
  )
  
  if (is.null(enrich_res) || nrow(as.data.frame(enrich_res)) == 0) {
    message("No significant enrichment found.")
    return(NULL)
  }
  
  # ---- Plot results
  print(
    dotplot(enrich_res, showCategory = showCategory) +
      ggtitle(paste(method, "Pathway Enrichment -", condition))
  )
  
  # ---- Convert to readable format (add gene symbols)
  readable_res <- suppressMessages(
    setReadable(enrich_res, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
  )
  
  result_df <- as.data.frame(readable_res)
  
  # ---- Save output to Excel
  output_file <- paste0(output_prefix, "_", condition, "_", method, "_pathway.xlsx")
  openxlsx::write.xlsx(result_df, output_file, overwrite = TRUE)
  message(paste("Saved:", output_file))
  
  return(result_df)
}
```

```{r}
# Run KEGG for R179C
R179C_KEGG<- Pathway_Enrich(sig_R179C_vs_WT, condition = "R179C", method = "KEGG")
R179C_KEGG

# Run Reactome for R179C
R179C_Reactome<- Pathway_Enrich(sig_R179C_vs_WT, condition = "R179C", method = "Reactome")
R179C_Reactome


```



## Pathway analysis 
```{r}
## R179C vs WT Pathway Analysis
library(clusterProfiler)
library(org.Mm.eg.db)
#library(enrichplot)

# Convert gene symbols to Entrez IDs
gene_symbols <- rownames(sig_R179C_vs_WT)
entrez_ids <- bitr(gene_symbols, fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Mm.eg.db)

## KEGG Pathway Analysis
if(!is.null(entrez_ids) && nrow(entrez_ids) > 0){
  kegg_enrich <- enrichKEGG(gene = entrez_ids$ENTREZID,
                            organism = 'mmu',
                            pvalueCutoff = 0.05)
  if(!is.null(kegg_enrich) && nrow(as.data.frame(kegg_enrich)) > 0){
    # Visualize KEGG results
    print(dotplot(kegg_enrich,
                  showCategory=10) +
            ggtitle("KEGG Pathway Enrichment"))
    # Convert KEGG pathway IDs to readable gene symbol 
    kegg_genes <- setReadable(kegg_enrich,
                                    OrgDb = org.Mm.eg.db,
                                    keyType = "ENTREZID")
    print(as.data.frame(kegg_genes))
    
  } else {
    print("No significant KEGG pathways found.")
  }
} else {
  print("No valid Entrez IDs found for KEGG analysis.")
}
```

return gene list that classified under each go term with all three category, Biological_Process, Molecular_Function, Cellular_Component
```{r}
library(openxlsx)
library(dplyr)
library(clusterProfiler)

export_top_GO_terms <- function(GO_result_list,
                                condition_name,
                                top_n = NULL,       # If NULL, export all dotplot-visible terms
                                output_prefix = "GO_Results",
                                remove_empty = TRUE # Option B: filter out terms without geneID
                                ) {
  
  ontologies <- c("Biological_Process", "Molecular_Function", "Cellular_Component")
  
  for (ont in ontologies) {
    
    if (!ont %in% names(GO_result_list)) {
      message(paste("Skipping", ont, "- not found in provided list."))
      next
    }
    
    go_obj <- GO_result_list[[ont]]$result
    go_expanded <- GO_result_list[[ont]]$expanded
    
    if (is.null(go_obj) || is.null(go_expanded)) {
      message(paste("Skipping", ont, "- missing data."))
      next
    }
    
    # convert enrichResult object to data frame
    go_df <- as.data.frame(go_obj)
    
    # ---- If top_n not specified, use same terms as dotplot (default = 10 or fewer)
    if (is.null(top_n)) {
      # Extract top categories shown in dotplot (based on p.adjust & gene ratio)
      top_terms <- go_df %>%
        arrange(p.adjust, desc(GeneRatio)) %>%
        slice_head(n = 10)
    } else {
      top_terms <- go_df %>%
        arrange(p.adjust) %>%
        slice_head(n = top_n)
    }
    
    filtered_expanded <- go_expanded %>%
      filter(ID %in% top_terms$ID)
    
    # ---- Option B: Remove terms with empty or NA geneID
    if (remove_empty) {
      filtered_expanded <- filtered_expanded %>%
        filter(!is.na(geneID), geneID != "")
    }
    
    # ---- Create Excel workbook
    wb <- createWorkbook()
    
    for (go_id in unique(filtered_expanded$ID)) {
      sub_df <- filtered_expanded %>% filter(ID == go_id)
      term_name <- unique(sub_df$Description)[1]
      
      # Clean & shorten Excel sheet name
      safe_name <- paste0(
        gsub("[^A-Za-z0-9]", "_", go_id), "_",
        gsub("[^A-Za-z0-9]", "_", substr(term_name, 1, 15))
      )
      safe_name <- substr(safe_name, 1, 31)
      
      addWorksheet(wb, safe_name)
      writeData(wb, safe_name, sub_df)
    }
    
    # ---- Save workbook per ontology
    filename <- paste0(output_prefix, "_", condition_name, "_", ont, "_filtered.xlsx")
    saveWorkbook(wb, filename, overwrite = TRUE)
    
    message(paste("saved:", filename))
  }
}

```

#### Add additional replicate to the data (4th replicate)
```{r}
library(dplyr)
library(purrr)

# 1. Define folder path - ensure this matches your folder name exactly
data_dir <- "../../bulkRNARefence1Replicate/Counts/"

# 2. Identify the count files
# Updated pattern to be more flexible (looks for 'counts' anywhere in the name)
count_files <- list.files(path = data_dir, 
                          pattern = "counts", 
                          full.names = TRUE)

# Filter out the 'Identifier' files manually to be safe
count_files <- count_files[!grepl("Identifier", count_files)]

# --- SAFETY CHECK ---
if (length(count_files) == 0) {
  stop("No files found! Check if './Count' is the correct folder and contains the .txt files.")
}

# 3. Read and Merge
counts_combined <- count_files %>%
  map(function(file) {
    # Using read.table; check if your files use tabs or commas
    df <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
    
    # Standardize the first column name for joining
    colnames(df)[1] <- "GeneID"
    
    # Extract a clean sample name from the file path
    sample_id <- basename(file) %>% gsub("_counts.*", "", .)
    colnames(df)[2] <- sample_id
    
    return(df)
  }) %>%
  reduce(full_join, by = "GeneID")

# 4. Final Formatting
rownames(counts_combined) <- counts_combined$GeneID
counts_combined <- as.matrix(counts_combined[,-1])

# View result
print(paste("Successfully merged", ncol(counts_combined), "files."))
#head(counts_combined)
```

## Identify Marker gene with Differential Expression 
```{r}
 sample_info_4rep<- data.frame(
  sampleName = c("R179C-1percent-1",
                 "R179C-1percent-2",
                 "R179C-1percent-3",
                 "R179C-1percent-4",
                 "WT-1percent-1",
                 "WT-1percent-2",
                 "WT-1percent-3",
                 "WT-1percent-4",
                 "WT-TGFb-1percent-1",
                 "WT-TGFb-1percent-2",
                 "WT-TGFb-1percent-3",
                 "WT-TGFb-1percent-4"),
  batch = c("Batch1","Batch1","Batch1","Batch2", 
                       "Batch1","Batch1","Batch1","Batch2",
                       "Batch1","Batch1","Batch1","Batch2"),
  condition = c("R179C", "R179C", "R179C", "R179C",
                "WT", "WT", "WT", "WT",
                "TGFb", "TGFb", "TGFb", "TGFb"),
  replicate = c("1", "2", "3", "4",
                "1", "2", "3", "4",
                "1", "2", "3", "4")
  
)
rownames(sample_info_4rep)<- sample_info_4rep$sampleName
```

## Replace Ensembl ID with gene symbol for the new replicate data
```{r}
# 1. Prepare counts_combined (Ensembl to Symbol Mapping)
# ---------------------------------------------------
# Strip version suffix (e.g., .15) from Ensembl IDs in the new replicate data
ens_ids <- sub("\\.\\d+$", "", rownames(counts_combined))
rownames(counts_combined) <- ens_ids

# Load your mapping file (ensure path is correct relative to your .Rmd)
mapfile <- "../../mm10_ens2symbol.tsv"
map <- read.delim(mapfile, header = FALSE, sep = "\t",
                  col.names = c("ensembl_gene_id", "gene_name"),
                  stringsAsFactors = FALSE)

# Remove duplicates in the map file to prevent 1-to-many expansion
map <- map[!duplicated(map$ensembl_gene_id), ]

# Create a lookup vector for quick translation
gene_lookup <- setNames(map$gene_name, map$ensembl_gene_id)

# Perform the mapping
matched_symbols <- gene_lookup[rownames(counts_combined)]

# If no symbol found in the TSV, keep the original Ensembl ID to avoid losing data
matched_symbols[is.na(matched_symbols)] <- rownames(counts_combined)[is.na(matched_symbols)]

# CRITICAL: Handle duplicate gene symbols (e.g., Symbol.1, Symbol.2)
# This prevents the "duplicate 'row.names' are not allowed" error in R matrices
rownames(counts_combined) <- make.unique(as.character(matched_symbols))
```


## Consistant rowname (gene id)
since there are only 5 common gene appear 
check the id by strip where everything after the dot would be remove 
this process won't actually affect the gene name 

```{r}
# Strip the version (anything after the first dot) from rownames
strip_ver <- function(x) sub("\\..*$", "", x)

ids_a <- strip_ver(rownames(count_matrix))
ids_b <- strip_ver(rownames(counts_combined))

length(intersect(ids_a, ids_b))


```
the intersection show that the common gene increase from 41 to 51756 
the next step is to apply it to the rownames of both datasets
```{r}
# Standardize rownames to versionless before merging
rownames(count_matrix)   <- ids_a
rownames(counts_combined) <- ids_b
colnames(counts_combined)<- c("R179C-1percent4", "WT-1percent4", "WT-TGFb-4")

```


```{r}
count_matrix_ordered <- merge(count_matrix, counts_combined, by = "row.names", all.x = TRUE)

# Clean up Row Names
rownames(count_matrix_ordered) <- count_matrix_ordered$Row.names
count_matrix_ordered$Row.names <- NULL

# Handle Missing Data (NAs)
# If a gene was in your original set but not the new one, we fill with 0
# This prevents DESeq2 from crashing due to missing values
count_matrix_ordered[is.na(count_matrix_ordered)] <- 0

# Final Formatting
count_matrix_ordered <- round(as.matrix(count_matrix_ordered))


desired_order <- c(
  "R179C-1percent-1_S21", "R179C-1percent-2_S24", "R179C-1percent-3_S27", "R179C-1percent4", # R179C Group
  "WT-1percent-1_S19", "WT-1percent-2_S22", "WT-1percent-3_S25", "WT-1percent4",             # WT Group
  "WT-TGFb-1percent-1_S20", "WT-TGFb-1percent-2_S23", "WT-TGFb-1percent-3_S26", "WT-TGFb-4" # TGFb Group
)

# 2. Reorder the Matrix columns
count_matrix_ordered <- count_matrix_ordered[, desired_order]
```


```{r}
dim(count_matrix_ordered)
```

```{r}
summary(count_matrix_ordered)
```


```{r}
hist(count_matrix_ordered, main= "Gene Expression", xlab = "Counts")
```
 
```{r}
colnames(count_matrix_ordered)<- rownames(sample_info_4rep)
dds_4rep<- DESeqDataSetFromMatrix(countData = count_matrix_ordered,
                             colData = sample_info_4rep,
                             design = ~ batch + condition)

dds_4rep$condition <- relevel(dds_4rep$condition, ref = "WT")

dds_4rep<- DESeq(dds_4rep)
```

creating pairwise analysis so we can compare the gene expression change between R179C vs WT and TGFb vs WT
```{r}
# pairwise
res_R179C_vs_WT_4rep   <- results(dds_4rep, contrast = c("condition","R179C","WT"))
```

## PCA and sample distance heatmap
```{r}
set.seed(123)

vsd_4rep <- vst(dds_4rep, blind = FALSE)
mat_4rep <- assay(vsd_4rep)

# This keeps all 8 replicates (4 vs 4) but ignores TGFb
vsd_4rep <- vsd_4rep[, vsd_4rep$condition %in% c("R179C", "WT")]
# Drop unused factor levels so the legend is clean
vsd_4rep$condition <- droplevels(vsd_4rep$condition)


# PCA with sample labels
library(ggplot2)
library(ggrepel) # You will need this for the labels

p <- plotPCA(vsd_4rep, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(p, "percentVar"))

ggplot(p, aes(PC1, PC2, color = condition, label = name)) +
  geom_point(size = 3) + 
  # Use ggrepel to prevent overlapping
  geom_text_repel(
    size = 3, 
    show.legend = FALSE,
    box.padding = 0.5,      # Add space around the text
    point.padding = 0.3,    # Add space between text and point
    max.overlaps = Inf      # Ensure all labels are shown
  ) +
  # Expand limits to give the labels room to breathe
  expand_limits(x = c(min(p$PC1) - 10, max(p$PC1) + 15)) +
  # Prevents labels from being cut off at the edge of the plot
  coord_cartesian(clip = "off") +
  xlab(paste0("PC1: ", percentVar[1], "%")) +
  ylab(paste0("PC2: ", percentVar[2], "%")) +
  ggtitle("PCA (vst) — all samples (4 Replicates)") +
  theme_bw() # Clean white background for better visibility

# Sample distance heatmap
d <- dist(t(assay(vsd_4rep)))
pheatmap(as.matrix(d), annotation_col=as.data.frame(colData(dds_4rep)[,"condition",drop=FALSE]))
```

```{r}
library(ggplot2)
library(ggrepel)
library(pheatmap)

# 1. Generate clean, sequential replicate numbers (1 to 4)
# This looks at the 'condition' and counts 1,2,3,4 for each group automatically
replicate_numbers <- ave(seq_along(vsd_4rep$condition), 
                         vsd_4rep$condition, 
                         FUN = seq_along)

# 2. Create the clean strings
new_names <- paste0(vsd_4rep$condition, "-", replicate_numbers)

# 3. Apply the new names to the object
colnames(vsd_4rep) <- new_names

# 4. Update PCA Data
p <- plotPCA(vsd_4rep, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(p, "percentVar"))

# PCA Plot with Clean Labels
ggplot(p, aes(PC1, PC2, color = condition, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(
    size = 4, 
    show.legend = FALSE,
    box.padding = 0.5, 
    point.padding = 0.3,
    fontface = "bold"
  ) +
  expand_limits(x = c(min(p$PC1) - 5, max(p$PC1) + 5)) +
  xlab(paste0("PC1: ", percentVar[1], "%")) +
  ylab(paste0("PC2: ", percentVar[2], "%")) +
  ggtitle("PCA: 4 Replicates per Condition (Cleaned)") +
  theme_bw(base_size = 14)

# 5. Update Sample Distance Heatmap
d <- dist(t(assay(vsd_4rep)))
pheatmap(as.matrix(d), 
         clustering_distance_rows = d,
         clustering_distance_cols = d,
         annotation_col = as.data.frame(colData(vsd_4rep)[,"condition",drop=FALSE]),
         main = "Sample Distance Heatmap (4 Replicates)",
         fontsize = 10)
```


## Check replicate 
since the fourth replicate is from another dataset 
the following chunk is designed to check if the batch effect got handle correctly 
```{r}
plotPCA(vsd_4rep, intgroup=c("condition","batch"))

```

```{r}
sample_info_clean_4rep<- sample_info_4rep %>% filter(condition %in% c("R179C", "WT"))
```


since the batch effect still exist check 
```{r}
library(limma)
expr_corrected <- removeBatchEffect(assay(vsd_4rep), batch=sample_info_clean_4rep$batch)

# PCA on corrected data
pca_corrected <- prcomp(t(expr_corrected))
plot(pca_corrected$x[,1:2], col=as.factor(sample_info_clean_4rep$condition), pch=19)
```

```{r}
# 1. Install/Load ggrepel if you haven't already
# install.packages("ggrepel")
library(ggplot2)
library(ggrepel)

# 2. Run PCA on corrected expression matrix
pca <- prcomp(t(expr_corrected)) 
pca_df <- as.data.frame(pca$x[, 1:2]) 
pca_df$sample <- rownames(pca_df)
pca_df$condition <- sample_info_clean_4rep$condition[match(pca_df$sample, sample_info_clean_4rep$sampleName)]

# 3. Calculate Variance Explained for Axis Labels
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)))

# 4. Plot with "Smart" Labels
ggplot(pca_df, aes(x=PC1, y=PC2, color=condition, label=sample)) +
  geom_point(size=4) + # Made points slightly larger for the poster
  # geom_text_repel is the fix for missing/merged labels
  geom_text_repel(
    size = 3.5,
    force = 2,            # Increases the "push" between overlapping labels
    box.padding = 0.5,    # Space around the text
    point.padding = 0.5,  # Space between point and text
    min.segment.length = 0, # Always draw a line from label to point if moved
    max.overlaps = Inf    # Forces all labels to show, even if crowded
  ) +
  # Expand the plot window so labels on the far edge aren't cut off
  scale_x_continuous(expand = expansion(mult = c(0.15, 0.2))) + 
  theme_bw() + # BW theme is cleaner for posters than minimal
  labs(
    title="PCA after Batch Correction (R179C vs WT)", 
    x=paste0("PC1: ", percentVar[1], "% variance"), 
    y=paste0("PC2: ", percentVar[2], "% variance")
  )
```

```{r}
library(ggplot2)
library(ggrepel)
library(dplyr)

# 1. Run PCA on corrected expression matrix
pca <- prcomp(t(expr_corrected)) 
pca_df <- as.data.frame(pca$x[, 1:2]) 
pca_df$sample <- rownames(pca_df)

# 2. MATCH AND CLEAN THE NAMES
# Get the condition from your metadata
pca_df$condition <- sample_info_clean_4rep$condition[match(pca_df$sample, sample_info_clean_4rep$sampleName)]

# Create the clean WT-1, R179C-1 labels
# This counts 1,2,3,4 based on the order they appear in the condition column
pca_df <- pca_df %>%
  group_by(condition) %>%
  mutate(clean_label = paste0(condition, "-", row_number())) %>%
  ungroup()

# 3. Calculate Variance
percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)))

# 4. Plot with Clean Labels
ggplot(pca_df, aes(x=PC1, y=PC2, color=condition, label=clean_label)) +
  geom_point(size=4) + 
  geom_text_repel(
    size = 4,
    force = 3,            # Increased force to push labels further for clarity
    box.padding = 0.6,    
    point.padding = 0.5,  
    fontface = "bold",    # Bold looks better on large posters
    min.segment.length = 0, 
    max.overlaps = Inf    
  ) +
  scale_color_manual(values = c("WT" = "#F8766D", "R179C" = "#00BFC4")) +
  # Expand limits so the labels "WT-4" or "R179C-4" don't hit the edge
  scale_x_continuous(expand = expansion(mult = c(0.2, 0.2))) + 
  theme_bw(base_size = 14) + 
  labs(
    title="PCA: Batch Corrected R179C vs WT", 
    x=paste0("PC1: ", percentVar[1], "% variance"), 
    y=paste0("PC2: ", percentVar[2], "% variance")
  )
```


## Heatmap of top variable genes after batch correction
```{r}
topVarGenes <- head(order(rowVars(expr_corrected), decreasing=TRUE), 500)

annotation <- data.frame(
  condition = sample_info_clean_4rep$condition,
  batch = sample_info_clean_4rep$batch
)
rownames(annotation) <- sample_info_clean_4rep$sampleName

pheatmap(expr_corrected[topVarGenes, ],
         annotation_col = annotation,
         show_rownames = FALSE,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")

```

```{r}
# 1. Define the specific flipped colors to match your PCA
# WT = Salmon (#F8766D) | R179C = Teal (#00BFC4)
# We leave 'batch' alone so it uses the default colors
ann_colors <- list(
  condition = c(WT = "#F8766D", R179C = "#00BFC4")
)

# 2. Ensure the annotation has all 8 samples
# Make sure sample_info_clean_4rep has 8 rows!
annotation <- data.frame(
  condition = sample_info_clean_4rep$condition,
  batch = as.factor(sample_info_clean_4rep$batch) # Factor ensures discrete colors
)
rownames(annotation) <- sample_info_clean_4rep$sampleName

# 3. Select Top 500 Genes
topVarGenes <- head(order(rowVars(expr_corrected), decreasing=TRUE), 500)

# 4. Generate Heatmap
library(pheatmap)
pheatmap(expr_corrected[topVarGenes, ],
         annotation_col = annotation,
         annotation_colors = ann_colors,   # <--- FLIPS THE GENOTYPE COLORS
         show_rownames = FALSE,
         show_colnames = FALSE,             
         cluster_rows = FALSE,             # Removes y-axis correlation as requested
         cluster_cols = TRUE,              # Keeps sample clustering
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")
```
```{r}
# 1. Clean the Sample Names (Remove "1-percent-")
# This creates clean "WT-1", "WT-2"... names for the x-axis
replicate_numbers <- ave(seq_along(sample_info_clean_4rep$condition), 
                         sample_info_clean_4rep$condition, 
                         FUN = seq_along)
clean_names <- paste0(sample_info_clean_4rep$condition, "-", replicate_numbers)

# 2. Update Dataframe and Matrix Names
annotation <- data.frame(
  condition = sample_info_clean_4rep$condition,
  batch = as.character(sample_info_clean_4rep$batch) # Character matches the list names below
)
rownames(annotation) <- clean_names

# Make a copy of the matrix to avoid changing your original data
plot_mat <- expr_corrected
colnames(plot_mat) <- clean_names

# 3. Define Consistent & High-Contrast Colors
# WT (Salmon) and R179C (Teal) to match PCA. Batch 2 is now Charcoal for contrast.
ann_colors <- list(
  condition = c(WT = "#F8766D", R179C = "#00BFC4"),
  batch = c("1" = "#D9D9D9", "2" = "#4D4D4D") 
)

# 4. Generate the Heatmap
library(pheatmap)
pheatmap(plot_mat[topVarGenes, ],
         annotation_col = annotation,
         annotation_colors = ann_colors,   # Flipped Genotype + Distinct Batch
         show_rownames = FALSE,
         show_colnames = TRUE,             # Now shows "WT-1", "R179C-1" etc.
         cluster_rows = FALSE,             # No y-axis correlation/dendrogram
         cluster_cols = TRUE,              # Keep sample clustering at top
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row",
         main = "Batch Corrected Transcriptional Profile (n=8)",
         color = colorRampPalette(c("royalblue", "white", "firebrick3"))(50),
         fontsize_col = 10)
```


```{r}
# 1. Ensure you have the corrected expression from earlier
library(limma)

# 2. Run Principal Component Analysis (PCA) on the corrected matrix
pca_corrected <- prcomp(t(expr_corrected))

# 3. Prepare data for ggplot
pca_df <- as.data.frame(pca_corrected$x)
pca_df$condition <- vsd_4rep$condition
pca_df$batch <- vsd_4rep$batch

# Calculate variance percentages for the axes
percentVar <- round(100 * (pca_corrected$sdev^2 / sum(pca_corrected$sdev^2)))

# 4. Plot the corrected version
ggplot(pca_df, aes(PC1, PC2, color = condition, shape = batch)) +
  geom_point(size = 4) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA: Batch Effect Removed (R179C vs WT)") +
  theme_bw()
```


## Heatmap of top variable genes after batch correction
```{r}
topVarGenes <- head(order(rowVars(expr_corrected), decreasing=TRUE), 500)

annotation <- data.frame(
  condition = sample_info_clean_4rep$condition,
  batch = sample_info_clean_4rep$batch
)
rownames(annotation) <- sample_info_clean_4rep$sampleName

pheatmap(expr_corrected[topVarGenes, ],
         annotation_col = annotation,
         show_rownames = FALSE,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "row")

```

### R179C vs wildType 
```{r}
sig_R179C_vs_WT_4rep<- subset(res_R179C_vs_WT_4rep, padj < 0.05
                         & abs(log2FoldChange) > 1)
sig_R179C_vs_WT_4rep<- as.data.frame(sig_R179C_vs_WT_4rep[order(sig_R179C_vs_WT_4rep$log2FoldChange, decreasing = T),])
nrow(sig_R179C_vs_WT_4rep)
head(sig_R179C_vs_WT_4rep, 10)
```

```{r}
write.csv(as.data.frame(sig_R179C_vs_WT_4rep), "./sig_R179C_vs_WT_4rep.csv")
```

## Create MA & volcano plot 
mark: previous simply calling plotMA likely calling plotMA from limma, 
      make sure to load the function from DESeq2 
#### R179C  vs WT
```{r}
DESeq2::plotMA(res_R179C_vs_WT_4rep, main = "R179C vs WT 4 replicate")
```

```{r}
library(EnhancedVolcano)

EnhancedVolcano(res_R179C_vs_WT_4rep,
                lab = rownames(res_R179C_vs_WT_4rep),   # gene labels
                x = 'log2FoldChange',              # x-axis: fold change
                y = 'padj',                        # y-axis: adjusted p-value
                title = 'R179C vs WT',             # plot title
                pCutoff = 0.05,                    # significance threshold
                FCcutoff = 1.0,                    # fold change threshold
                pointSize = 2.0,                   # point size
                labSize = 3.0)                     # label size
```

## plot count example 
use Prune2 gene as example
```{r}
# 1. Subset the dds object to the 8 samples (4 vs 4)
# This removes the TGFb samples from the counts and metadata
dds_subset_4rep <- dds_4rep[, dds_4rep$condition %in% c("R179C", "WT")]

# 2. Drop the 'WT-TGFb' level so it doesn't appear in your plots
dds_subset_4rep$condition <- droplevels(dds_subset_4rep$condition)

# 3. CRITICAL: Re-run the DESeq estimation
# This ensures the size factors (normalization) are based ONLY on these 8 samples
dds_subset_4rep <- DESeq(dds_subset_4rep)

# 4. Plot individual gene counts (e.g., Prune2)
plotCounts(dds_subset_4rep, gene = "Prune2", intgroup = "condition", main = "Prune2: R179C vs WT")
```

## check if the gene list shown in-consistancy in mutant with stable WT expression 
```{r}
geneList<- c("Acta2", "Cnn1", "Tagln", "Myh11", "Tspan2", 
                  "Itga8", "Smtn", "Dhx32os", "Myocd", "Mrtfa", "Lmod1")
geneList
#geneList %in% rownames(count_matrix)
gene_valid<- geneList[geneList %in% rownames(count_matrix)]
gene_valid

for(i in 1:length(gene_valid)){
  plotCounts(dds_subset, gene = gene_valid[i], intgroup = "condition", main = paste0(gene_valid[i], ": R179C vs WT"))
}
```

#### R179C vs WT separated for upregulated and downregulated genes
```{r}
# R179C upregulated genes
#r179c_up <- subset(res_R179C_vs_WT, log2FoldChange > 1 & padj < 0.05)
#top20_R179C <- head(rownames(r179c_up[order(r179c_up$padj), ]), 20)
top20_R179C_4rep <- head(rownames(sig_R179C_vs_WT_4rep), 20)

# WT upregulated genes
#wt_up <- subset(res_R179C_vs_WT, log2FoldChange < -1  & padj < 0.05)
#top20_WT <- head(rownames(wt_up[order(wt_up$padj), ]), 20)

# Already filtered for padj < 0.05 and |log2FC| > 1
sig_R179C_vs_WT_WT_4rep <- sig_R179C_vs_WT_4rep[order(sig_R179C_vs_WT_4rep$log2FoldChange), ]
top20_WT_4rep <- head(rownames(sig_R179C_vs_WT_WT_4rep), 20)

# Combine gene set
combined_genes <- c(top20_R179C_4rep, top20_WT_4rep)

# Subset samples to R179C and WT only 
selected_samples <- colData(dds_4rep)$condition %in% c("R179C", "WT")

mat_combined <- assay(vst(dds_4rep))[combined_genes, selected_samples]
mat_combined <- mat_combined - rowMeans(mat_combined)

annotation <- as.data.frame(colData(dds_4rep)[selected_samples, "condition", drop=FALSE])

pheatmap(mat_combined,
         annotation_col = annotation,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE, 
         main = "Heatmap of Top 20 DE Genes: R179C vs WT")
```


## pathway analysis function
```{r}
suppressMessages({
  library(clusterProfiler)
  library(ReactomePA)
  library(org.Mm.eg.db)
  library(ggplot2)
  library(openxlsx)
})

Pathway_Enrich <- function(sig_gene_list,
                           condition = "Sample",
                           method = c("KEGG", "Reactome"),
                           p_cut = 0.05,
                           showCategory = 10,
                           output_prefix = "Pathway") {
  
  method <- match.arg(method)  # restrict to KEGG or Reactome
  
  message(paste0("Running ", method, " enrichment for ", condition, " ..."))
  
  # ---- Convert gene symbols to Entrez IDs
  gene_symbols <- rownames(sig_gene_list)
  entrez_ids <- suppressMessages(
    bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
  )
  
  if (is.null(entrez_ids) || nrow(entrez_ids) == 0) {
    warning("No valid Entrez IDs found for enrichment.")
    return(NULL)
  }
  
  # ---- Run enrichment analysis
  enrich_res <- switch(
    method,
    KEGG = suppressMessages(
      enrichKEGG(gene = entrez_ids$ENTREZID,
                 organism = "mmu",
                 pvalueCutoff = p_cut)
    ),
    Reactome = suppressMessages(
      enrichPathway(gene = entrez_ids$ENTREZID,
                    organism = "mouse",
                    pvalueCutoff = p_cut)
    )
  )
  
  if (is.null(enrich_res) || nrow(as.data.frame(enrich_res)) == 0) {
    message("No significant enrichment found.")
    return(NULL)
  }
  
  # ---- Plot results
  print(
    dotplot(enrich_res, showCategory = showCategory) +
      ggtitle(paste(method, "Pathway Enrichment -", condition))
  )
  
  # ---- Convert to readable format (add gene symbols)
  readable_res <- suppressMessages(
    setReadable(enrich_res, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
  )
  
  result_df <- as.data.frame(readable_res)
  
  # ---- Save output to Excel
  output_file <- paste0(output_prefix, "_", condition, "_", method, "_pathway.xlsx")
  openxlsx::write.xlsx(result_df, output_file, overwrite = TRUE)
  message(paste("Saved:", output_file))
  
  return(result_df)
}
```

```{r}
# Run KEGG for R179C
R179C_KEGG_4rep<- Pathway_Enrich(sig_R179C_vs_WT_4rep, condition = "R179C", method = "KEGG")
R179C_KEGG_4rep

# Run Reactome for R179C
R179C_Reactome_4rep<- Pathway_Enrich(sig_R179C_vs_WT_4rep, condition = "R179C", method = "Reactome")
R179C_Reactome_4rep

```

## Pathway analysis for GO Pathway analysis 
## Function convert enrichment result to readable gene symbols
```{r}
library(clusterProfiler)
library(org.Mm.eg.db)

covert_entrez_to_symbols<- function(enrichment_result){
  if(!is.null(enrichment_result) &&
     nrow(enrichment_result)> 0){
    enrichment_df<- as.data.frame(enrichment_result)
    
    #Conver Entrez IDs to gene Symbols 
    enrichment_df$GeneSymvols<-
      sapply(strsplit(enrichment_df$geneID, "/"), function(entrez_list){
       mapped_genes<- bitr(entrez_list, fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Mm.eg.db)
       paste(mapped_genes$SYMBOL, collapse = ", ")
      })
    return(enrichment_df)
  } else {
    return(NULL)
  }
  
}
```

### go term for portable gene list
Biological Process (BP): This domain encompasses series of molecular events or functions, representing larger biological objectives that are accomplished by the concerted action of multiple molecular activities.

Molecular Function (MF): This domain describes the activities of individual gene products at the molecular level, such as catalyzing chemical reactions or binding specific chemicals or macromolecules.

Cellular Component (CC): This domain refers to the parts of a cell, including subcellular structures, macromolecular complexes, and the extracellular environment where gene products are located.
```{r}
GO_Res <- function(sig_gene_list,
                   p_cut = 0.05,
                   organism = "mouse") {
  # --- Step 1: Load OrgDb based on organism
  if (organism == "mouse") {
    OrgDb <- org.Mm.eg.db
  } else if (organism == "human") {
    OrgDb <- org.Hs.eg.db
  } else {
    stop("Unsupported organism. Use 'mouse' or 'human'.")
  }

  # --- Step 2: Prepare gene IDs
  gene_symbols <- rownames(sig_gene_list)
  entrez_ids <- bitr(gene_symbols,
                     fromType = "SYMBOL",
                     toType = "ENTREZID",
                     OrgDb = OrgDb)
  
  # Filter valid Entrez IDs only
  entrez_ids <- entrez_ids[!is.na(entrez_ids$ENTREZID) & entrez_ids$ENTREZID != "", ]
  
  if (is.null(sig_gene_list) || nrow(entrez_ids) == 0) {
    message("No valid gene list or unsupported organism.")
    return(NULL)
  }
  
  # --- Step 3: Define helper function for enrichment and table generation
  run_GO <- function(ont) {
    result <- enrichGO(gene = entrez_ids$ENTREZID,
                       OrgDb = OrgDb,
                       ont = ont,
                       pvalueCutoff = p_cut,
                       readable = TRUE)
    if (is.null(result) || nrow(as.data.frame(result)) == 0) {
      return(list(result = NULL, df = NULL, expanded = NULL))
    }
    
    df <- as.data.frame(result)
    
    # Expand geneID column into individual genes
    expanded <- df %>%
      dplyr::select(ID, Description, geneID) %>%
      tidyr::separate_rows(geneID, sep = "/") %>%
      dplyr::distinct(geneID, .keep_all = TRUE)
    
    # Print summary
    print(dotplot(result) + ggtitle(paste("GO", ont, "Enrichment")))
    print(df)
    
    return(list(result = result, df = df, expanded = expanded))
  }

  # --- Step 4: Run GO enrichment for BP, MF, CC
  go_bp <- run_GO("BP")
  go_mf <- run_GO("MF")
  go_cc <- run_GO("CC")
  
  # --- Step 5: Combine into a nested return list
  GO_out <- list(
    Biological_Process = go_bp,
    Molecular_Function = go_mf,
    Cellular_Component = go_cc
  )
  
  message("GO analysis completed successfully.")
  return(GO_out)
}

```

#### R179C vs WT GO Pathway Analysis with expanded gene list
```{r}
GO_R179C_vs_WT_4rep <- suppressMessages(
   GO_Res(sig_gene_list = sig_R179C_vs_WT_4rep)
)
```


```{r}
commPathWayFunc<- function(gene,
                           method = "KEGG",
                           go_type = NULL){
  
  entrez <- bitr(gene, fromType =  "SYMBOL",
                 toType = "ENTREZID", OrgDb = org.Mm.eg.db)
  
  if(method == "KEGG"){
    kegg<- enrichKEGG(gene = entrez$ENTREZID,
               organism = "mmu",
               pvalueCutoff = 0.05)
    return(kegg)
  } else if (method == "Reactome"){
    reactome<- enrichPathway(gene = entrez$ENTREZID,
                  organism = "mouse",
                  pvalueCutoff = 0.05)
    return(reactome)
  } else if (method == "GO" & !is.null(go_type)){
    go<- enrichGO(gene = entrez$ENTREZID,
             OrgDb = org.Mm.eg.db,
             ont = go_type,
             pvalueCutoff = 0.05,
             readable = TRUE)
    return(go)
  }
  
}
```

#### Common gene between R179C 3rep and 4rep
```{r}
dim(sig_R179C_vs_WT)
dim(sig_R179C_vs_WT_4rep)
common_R179C_genes <- intersect(rownames(sig_R179C_vs_WT), rownames(sig_R179C_vs_WT_4rep))
unique_3rep<- setdiff(rownames(sig_R179C_vs_WT), rownames(sig_R179C_vs_WT_4rep))

length(unique_3rep)
unique_4rep<- setdiff(rownames(sig_R179C_vs_WT_4rep), rownames(sig_R179C_vs_WT))
length(unique_4rep)

length(common_R179C_genes)
```

```{r}
sig_R179C_vs_WT_common<- sig_R179C_vs_WT_4rep[common_R179C_genes,]
```

```{r}
#install.packages("ggvenn")

library(ggvenn)

ggvenn(
  list(
    `3rep` = rownames(sig_R179C_vs_WT),
    `4rep` = rownames(sig_R179C_vs_WT_4rep)
  ),
  fill_color = c("skyblue", "salmon"),
  stroke_size = 0.5,
  set_name_size = 5
)
```


#### compare R179C KEGG results 
```{r}
cc_kegg<- function(gene){
  commPathWayFunc(gene, method = "KEGG")
}

compareCluster(geneClusters = list(
  "3rep" = unique_3rep,
  "4rep" = unique_4rep,
  "common" = common_R179C_genes
), fun = cc_kegg) %>%
  dotplot(showCategory=10) + ggtitle("KEGG Pathway #Comparison: R179C 3rep vs 4rep")

```



